---
title: "phenotype_prediction_plots.Rmd"
author: "Eva"
date: "2023-06-20"
output: html_document
---

```{r}
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(cowplot)
library(stringr)
library(arrow)
library(yardstick)
library(gdata)
library(dplyr)
library(tidyr)
library(purrr)
library(latex2exp)

options(dplyr.summarise.inform = FALSE)

```

```{r}
experiment_directory = 'PATH_TO_EXPERIMENT" # where deeprvat-analysis/phenotype_prediction/phenotype_prediction_pipeline/phenotype_prediction_pipeline.snakefile has been run' #Has to be put in!

plot_out_path = "./plots"
```

```{r}

phenotypes = c("Apolipoprotein_A", 
               "Apolipoprotein_B", 
               "Calcium", 
               "Cholesterol",
               "HDL_cholesterol",
               "IGF_1" ,
               "LDL_direct",
               "Lymphocyte_percentage",
               "Mean_corpuscular_volume",
               "Mean_platelet_thrombocyte_volume",
               "Mean_reticulocyte_volume",
               "Neutrophill_count",
               "Platelet_count",
               "Platelet_crit",
               "Platelet_distribution_width",
               "Red_blood_cell_erythrocyte_count",
               "SHBG",
               "Standing_height",
               "Triglycerides",
               "Total_bilirubin",
               "Urate"
)

phenotype_renamer = c(
    'Mean_platelet_thrombocyte_volume' = 'MPTVS',
    'IGF_1' = 'IGF-1',
    'Red_blood_cell_erythrocyte_count' = 'Erythrocyte count'
)

phenotype_dict = c()
for (p in phenotypes){
  if (! p %in% names(phenotype_renamer)){
    p_name = str_replace_all(p, '_', ' ')
  }
  else{
    p_name = phenotype_renamer[[p]]
  }
  phenotype_dict[[p]] = p_name
}

paper_plots = list()
```



```{r}
getwd()
linear_model_res_path = file.path(experiment_directory,'/phenotype_prediction/models/linear_models/')
fdr = 0.05
phenotype_suffix = 'qt'
phenotypes <- phenotypes[file.exists(paste0(linear_model_res_path, '/', phenotypes, '/', phenotype_suffix, '_fdr-', fdr, '.Rds'))]
gene_list_renamer = c('Burden/SKAT combined', 'DeepRVAT discoveries')
names(gene_list_renamer) = c('baseline_only', 'deeprvat_discoveries')

```

```{r}
burdens_to_exclude = c('cadd_10', 'cadd_20', 'polyphen_damaging', 'absplice')
paste(burdens_to_exclude, collapse = '|')
all_combined_metrics = readRDS(file.path(linear_model_res_path, 'plotting_data', paste0('all_recomputed_metrics_test_', phenotype_suffix, '_', fdr,'.Rds'))) %>%
  filter(!grepl(paste(burdens_to_exclude, collapse = '|'), model_name))
all_combined_metrics %>% distinct(model_name)
```

```{r}
baseline_btypes = unique(sapply(strsplit(grep('baseline-', all_combined_metrics %>% distinct(model_name) %>% pull(model_name), value = TRUE), "-"), function(x) x[2]))

baseline_methods =  paste('baseline', baseline_btypes, sep = '-')
method_gene_list_filter = quo((`rare_burden` %in% baseline_methods & gene_list == 'baseline_only') | (`rare_burden` == 'deeprvat-avg' & gene_list == 'deeprvat_discoveries'))
method_gene_list_filter
```


## Plotting 
```{r}
label_default_size = 14 #default ggarrange size
default_plot_width = 15.5 #used by brian


method_names = c(#'baseline-cadd_10' = 'CADD >10', 
                 #'baseline-cadd_20' = 'CADD >20', 
                  "baseline-cadd" = 'CADD',
                 "baseline-plof" = 'pLOF',
                 #"baseline-polyphen_damaging" = 'PolyPhen2 Probably Damaging',
                 "baseline-polyphen" = 'PolyPhen2',
                "baseline-primateai" = 'PrimateAI',
                "baseline-sift"= 'SIFT',
                "baseline-splicai" = 'SpliceAI',
                # 'baseline-absplice'= 'AbSplice',
                'deeprvat-avg' = 'DeepRVAT',
                'covariates-PRS' = 'common PRS')

viridis_colors = viridis::viridis(n = length(baseline_methods) + 4)
deeprvat_color = '#7A2961'
common_prs_color = 'darkgoldenrod2'



RmUnderscore <- function(string) {
  sub("_", " ", string)
}
remove_x_underscore = scale_x_discrete(labels = ~ gsub("_", " ", .x)) 
remove_y_expand = scale_y_continuous(labels = scales::percent, expand =  expansion(mult = c(0, 0.1)))

facet_background = strip.background = element_rect(fill = "white", colour = "black", 
                linewidth = rel(2))
```

# Load and prepare data

## linear model results
```{r}
model_order = c("covariates", "covariates-PRS", baseline_methods, "deeprvat-avg") #, "deeprvat-0","deeprvat-1","deeprvat-2","deeprvat-3","deeprvat-4","deeprvat-5")
model_order

all_combined_metrics = all_combined_metrics %>% mutate(gene_list = ifelse(grepl('covariates', model_name), 'None', str_split(model_name, '-',  simplify = TRUE)[,3]))

# deeprvat_avg = all_combined_metrics %>% filter(method == 'deeprvat') %>% 
#   group_by_at(vars(-model_name, -.estimate)) %>%
#   summarise(.estimate = mean(.estimate)) %>% mutate(model_name = paste('deeprvat-avg', gene_list, sep = '-'))
# 
# 
# all_combined_metrics = rbind(all_combined_metrics, deeprvat_avg)

#dummy rows for covariates for all gene lists
covariates_res = all_combined_metrics %>% filter(grepl('covariates', model_name))
for (this_gene_list in unique(all_combined_metrics$gene_list)){
  all_combined_metrics = rbind(all_combined_metrics, covariates_res %>% mutate(gene_list = this_gene_list))
}
covariates_res

all_combined_metrics =  all_combined_metrics %>% 
  separate(model_name, c('one', 'two', 'three'), sep = '-', remove = FALSE) %>%
  mutate(model = paste(one, two, sep = '-')) %>% select(-one, -two, -three, -method) %>%
  mutate(model = str_replace(model, "covariates-NA", "covariates")) %>%
  filter(!grepl('None', gene_list))  %>%
  select(-top_quantile) %>%
  filter(model %in% model_order)
all_combined_metrics[['model']] =  factor(all_combined_metrics[['model']], model_order)
# all_combined_metrics[['model_name_plot']] = method_names[all_combined_metrics[['model']]]
```


```{r}
unique(all_combined_metrics$model)
delta_df = all_combined_metrics %>%
  # filter(model %in% c("baseline-plof", "baseline-cadd",
  #                                                       "covariates-PRS", "deeprvat-avg")) %>%
  select(-model_name) %>%
  # filter(gene_list == "deeprvat_discoveries" & .metric == 'rsq') %>%
  pivot_wider(names_from = model, values_from = .estimate) %>%
  pivot_longer(all_of(grep('covariates', model_order, invert = TRUE, value = TRUE)),
               names_to = 'rare_burden', values_to = 'r2_rare') %>%
  mutate(delta_r2 = r2_rare - `covariates-PRS`) %>%
  mutate(ratio_r2 = delta_r2/`covariates-PRS` )
delta_df %>% filter(.metric == 'rsq')


```

```{r}
method_map =  c('topq' = 'Top', 'bottomq' = 'Bottom', 'both' = 'highest/largest')
```

## logistic model results
```{r}
logistic_model_res_path = file.path(experiment_directory, 'phenotype_prediction/models/logistic_models/')
logistic_model_metrics = readRDS(file.path(logistic_model_res_path, 'plotting_data', paste0('combined_metrics_', phenotype_suffix, '_', fdr,'.Rds')))

logistic_model_metrics = logistic_model_metrics %>% 
  mutate(model_name = sub('deeprvat-', 'deeprvat-avg-', model_name)) %>%
  separate(col = model_name, into = c('burden', 'vtype_repeat', 'gene_list'), sep = '-', remove = FALSE) %>%
  filter(!grepl('_', vtype_repeat)) %>% #remove cadd_10 ..
  mutate(rare_burden = paste(burden, vtype_repeat, sep = '-'))
logistic_model_metrics %>% distinct(vtype_repeat)
logistic_model_metrics %>% distinct(rare_burden)


covariates_res = logistic_model_metrics %>% filter(grepl('covariates', model_name))
for (this_gene_list in na.omit(unique(logistic_model_metrics$gene_list))){
  logistic_model_metrics = rbind(logistic_model_metrics, covariates_res %>% mutate(gene_list = this_gene_list))
}
logistic_model_metrics = drop_na(logistic_model_metrics, gene_list)
rare_burden_models = grep('covariates', unique(logistic_model_metrics$rare_burden), invert = TRUE, value = TRUE)

delta_df_logistic = logistic_model_metrics %>% select( -model_name, -vtype_repeat, -burden) %>%
  pivot_wider(names_from = rare_burden, values_from = .estimate) %>% 
  select(-`covariates-NA`) %>%
  pivot_longer(rare_burden_models,
               names_to = 'rare_burden', values_to = 'auprc_rare') %>%
  mutate(delta_auprc = auprc_rare - `covariates-PRS`,
         ratio_delta_auprc = (auprc_rare - `covariates-PRS`)/ `covariates-PRS`)

delta_df_logistic %>% distinct(rare_burden, gene_list)

delta_df_logistic = delta_df_logistic %>% mutate(extreme_quantile_str = paste0(top_quantile * 100, '%'))
delta_df_logistic[['extreme_direction_string']] = method_map[as.character(delta_df_logistic[['extreme_direction']])]
delta_df_logistic = delta_df_logistic %>% mutate(outlier_mode_name = paste(extreme_direction_string, extreme_quantile_str))

```

## Replication in extreme quantiles plot

data generated in in genopheno/phenotype_prediction/replication_in_extremes_plot.R
```{r}
extreme_quantiles = c(0.01)
ranked_df_list = list()
for (outlier_mode in c('topq', 'bottomq')){
  for (extreme_quantile in extreme_quantiles){
      if (outlier_mode == 'both'){
        #here in total extreme_quantile *2 percent have the data been used (exreme_quantile % at each extreme)
        file_path = file.path(linear_model_res_path, 'plotting_data', 'replication_in_extreme', paste0('ranked_df_combined_', 
                                                    outlier_mode, '_', extreme_quantile, '_x2_', phenotype_suffix, '_', fdr,'.Rds'))
      }else{
        file_path = file.path(linear_model_res_path, 'plotting_data', 'replication_in_extreme', paste0('ranked_df_combined_', 
                                            outlier_mode, '_', extreme_quantile, '_', phenotype_suffix, '_', fdr,'.Rds'))
      }
      ranked_df = readRDS(file_path)
      ranked_df_list = append(ranked_df_list, list(ranked_df))
  }
}

ranked_df_combined = do.call(rbind, ranked_df_list) %>%
  mutate(extreme_quantile_str = paste0(extreme_quantile * 100, '%'))
ranked_df_combined[['outlier_mode']] = factor(method_map[ranked_df_combined[['outlier_mode']]], levels = method_map)
# mean(ranked_df_combined %>% filter(max_rank == 21 & `Rare estimator` == 'baseline-cadd') %>% pull(sum))

ranked_df_combined_avg = ranked_df_combined %>% group_by(`Rare estimator`, max_rank, extreme_quantile, extreme_quantile_str, outlier_mode) %>%
  summarise(sum = mean(sum)) %>% mutate(phenotype = 'All Traits') %>%
  mutate(outlier_mode_name = paste(outlier_mode, extreme_quantile_str))
ranked_df_combined_avg
```


run genopheno/phenotype_prediction/plot_enrichment_prs_vs_szsore.R
```{r}

enrichment_list = list()
for (this_quantile in c(0.99, 0.999)){
  this_enrichment = readRDS(file.path(linear_model_res_path, 'plotting_data', paste0('enrichment_prs_vs_zscore_', 
                                                                                   this_quantile, '_', phenotype_suffix, '_', fdr,'.Rds')))
  enrichment_list = append(enrichment_list, list(this_enrichment))
}
enrichment_combined = do.call(rbind, enrichment_list)


```


## enrichment in extreme phenotypes

```{r}
zero_refernce = enrichment_combined %>% filter(min_z_score == 0) %>%
  mutate(prop_zero = extremes/size) %>%
  select(-size, -extremes, -min_z_score)

enrich_score_df = enrichment_combined %>% 
  filter(min_z_score > 0) %>%
  merge(zero_refernce, how = 'left') %>%
  mutate(prop_extreme_samples = extremes/size,
         enrichment = prop_extreme_samples/prop_zero) %>%
  mutate(prs_quantile = paste0(prs_quantile * 100, '% quantile'))

enrich_score_df[!complete.cases(enrich_score_df), ]


avg_rel_improvement_df = enrich_score_df  %>% select(model, prs_quantile, min_z_score, enrichment, phenotype) %>%
  pivot_wider(names_from = model, values_from = enrichment) %>% 
  pivot_longer(cols = c(starts_with('baseline-'), 'deeprvat-avg'), names_to = 'model') %>%
  # filter(`covariates-PRS` > 0 & value > 0) %>%
  mutate(improvment_common = value - `covariates-PRS`, rel_improvment_common = improvment_common/`covariates-PRS`) %>%
  group_by(model, prs_quantile, min_z_score) %>%
  summarise(rel_improvment_common = mean(rel_improvment_common), improvment_common = mean(improvment_common)) %>%
  # summarise(enrichment = mean(prop_extreme_samples)/mean(prop_zero)) %>%
  ungroup()


enrich_score_avg_df = enrich_score_df %>% #%>% filter(phenotype != 'Standing_height') %>%
  group_by(model_name, model, burden, vtype_repeat,  prs_quantile, min_z_score) %>%
  summarise(enrichment = mean(enrichment)) %>%
  # summarise(enrichment = mean(prop_extreme_samples)/mean(prop_zero)) %>%
  ungroup()
```

### get largest threshold where enrichment for no phenotype/method is 0
```{r}
threshold_range <- seq(3, 4, by = 0.01)

# Initialize the max_z_score variable
max_z_score <- 0

# Iterate over each threshold in the range
for (threshold in threshold_range) {
  # Filter the data frame based on the threshold
  filtered_df <- enrich_score_df[enrich_score_df$min_z_score <= threshold, ]
  
  # Check if all values in Value2 are non-zero
  if (all(filtered_df$enrichment != 0)) {
    max_z_score <- threshold
  }
}

# Print the largest threshold at which no value in Value2 is zero
print(max_z_score)
```

## Main figures

```{r}
## determine boxplot order (lowest to highest median from linear models)
method_order = delta_df %>%
             filter(fdr == 0.05 & .metric == 'rsq') %>%
               filter(!! method_gene_list_filter) %>%
  group_by(rare_burden) %>%
  summarise(med = median(ratio_r2)) %>% arrange(med) %>%
  pull(rare_burden)

color_mapping = c(tail(viridis_colors, -4), deeprvat_color, common_prs_color)
names(color_mapping) = c(names(method_names[method_order]), 'covariates-PRS')

color_mapping

color_mapping_sub = color_mapping[grep('_', names(color_mapping), invert = TRUE)]

rare_method_fill = scale_fill_manual(values = color_mapping)
rare_method_color = scale_fill_manual(values = color_mapping)
```

```{r}
# mean and median imporovement
delta_df %>%
             filter(fdr == 0.05 & .metric == 'rsq') %>%
               filter(!! method_gene_list_filter) %>%
  group_by(rare_burden) %>%
  summarise(median = median(ratio_r2)*100, mean = mean(ratio_r2)*100)
```


```{r}
paper_plots = list()
```


```{r}
my_comparisons <- list()
for (method in baseline_methods) {
  my_comparisons <- c(my_comparisons, list(c(method, "deeprvat-avg")))
}


max_val = quantile(delta_df %>%
             filter(fdr == 0.05 & .metric == 'rsq') %>%
               filter(!! method_gene_list_filter) %>% pull(ratio_r2), 0.9999)
n_items = round((length(unique(delta_df$rare_burden)) -1)/2,0)
pos_sig = seq(from = max_val, by = 0.003, length.out = n_items)
pos_sig_2 = seq(from = max_val + 0.0015 , by = 0.003, length.out = length(my_comparisons) - n_items)



p_box_lin = ggplot(data=delta_df %>%
             filter(fdr == 0.05 & .metric == 'rsq') %>%
               filter(!! method_gene_list_filter),
           aes(y=ratio_r2, x = rare_burden, fill = rare_burden)) +
          geom_boxplot() +
          # rare_method_fill +
          labs( y= expression(paste("Relative \u2206", R^2)), fill = 'Rare variant \nscoring method', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = c(head(baseline_methods, n_items), 'deeprvat-avg', tail(baseline_methods, -n_items)),
                             labels = method_names) +
          scale_fill_manual(values = color_mapping, limits = names(color_mapping),labels = names(color_mapping), guide = 'none') + #limits = c(baseline_methods, 'deeprvat-avg')
          scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + #limits = c(baseline_methods, 'deeprvat-avg') +
          theme_cowplot() +
          geom_signif(comparisons = my_comparisons, color = 'grey30',
                      map_signif_level=c("***"=0.001, "**"=0.01, "*"=0.05), 
                      test = "wilcox.test",
                      test.args = c(alternative = 'less'),
                      y_position = c(pos_sig, pos_sig_2), 
                      # x_min()
                      # step_increase = 0.1,
                      size = 0.2, vjust = 0.5,
                      tip_length = 0.02) +
          remove_y_expand +
          theme(axis.text.x = element_text(angle = 44,  hjust =1))
print(p_box_lin)
#ggsave(file.path(plot_out_path,'linear_box_rsquared.png'), p_box_lin, height = 6, width = 4)

# paper_plots[['linear_boxplot']] = p_box_lin
```
```{r}
# only inidicate if DeepRVAT  is not significantly larger than others 
p_box_lin_ns_mark_only = ggplot(data=delta_df %>%
             filter(fdr == 0.05 & .metric == 'rsq') %>%
               filter(!! method_gene_list_filter),
           aes(y=ratio_r2, x = rare_burden, fill = rare_burden)) +
          geom_boxplot() +
          # rare_method_fill +
          labs( y= expression(paste("Relative \u2206", R^2)), fill = 'Rare variant \nscoring method', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = method_order,
                             labels = method_names) +
          scale_fill_manual(values = color_mapping, limits = names(color_mapping),labels = names(color_mapping), guide = 'none') +
          scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') +
          theme_cowplot() +
          ##
          # everything is siginificant so no comparison needed
          ##
          # geom_signif(comparisons = my_comparisons, color = 'grey30',
          #             map_signif_level=c("***"=0.001, "**"=0.01, "*"=0.05), 
          #             test = "wilcox.test",
          #             test.args = c(alternative = 'less'),
          #             y_position = c(pos_sig, pos_sig_2), 
          #             # x_min()
          #             # step_increase = 0.1,
          #             size = 0.2, vjust = 0.5,
          #             tip_length = 0.02) +
          remove_y_expand +
          theme(axis.text.x = element_text(angle = 44,  hjust =1))
print(p_box_lin_ns_mark_only)

paper_plots[['linear_boxplot']] = p_box_lin_ns_mark_only
```


```{r}
metric = 'ratio_delta_auprc'
this_top_quantile = 0.01
n = 0
p_box_list = list()
for (top_bottom_q in c('bottomq', 'topq')){

  plot_df = delta_df_logistic %>%
     filter(extreme_direction == top_bottom_q & top_quantile == this_top_quantile) %>%
         filter(fdr == 0.05 & .metric == 'pr_auc') %>%
           filter(!! method_gene_list_filter)
  
  y_min = min(plot_df[[metric]])
  y_max = quantile(plot_df[[metric]], 0.9999) #excluding extreme oultiers
  y_max = max(plot_df[[metric]][plot_df[[metric]] < y_max])
  pos_sig = seq(from = y_max, by = 0.03, length.out = n_items)
  pos_sig_2 = seq(from = y_max + 0.01 , by =  0.03, length.out = length(my_comparisons) - n_items)
  
  p_box = ggplot(#data=plot_df[plot_df[[metric]] <= y_max,],
                 data= plot_df,
             aes_string(y = metric, x = 'rare_burden', fill = 'rare_burden')) +
            geom_boxplot() +
            geom_hline(yintercept = 0, color = 'grey60') +
            facet_wrap(vars(outlier_mode_name), nrow = 1) +
            labs( y= expression(paste("Relative \u2206AUPRC")), 
                  fill = 'Rare variant \nscoring method', x = '') + # 'Relative  \u2206  R2'
            scale_x_discrete(limits = c(head(baseline_methods, n_items), 'deeprvat-avg', tail(baseline_methods, -n_items)),
                             labels = method_names) +
            scale_fill_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + 
            scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + 
            theme_cowplot() +
            geom_signif(data = plot_df, color = 'grey30',
                        comparisons = my_comparisons, 
                        map_signif_level=c("***"=0.001, "**"=0.01, "*"=0.05, "ns" = 0.51), 
                        test = "wilcox.test",
                        test.args = c(alternative = 'less'),
                        y_position = c(pos_sig, pos_sig_2), 
                        size = 0.2, vjust = 0.5,
                        tip_length = 0.01) +
            scale_y_continuous(labels = scales::percent) +
           theme(axis.text.x = element_text(angle = 44,  hjust =1),
                         strip.background = facet_background)
  if (n > 0){
    print('removing title')
    p_box = p_box + theme(axis.title.y = element_blank())
  }
  n = n + 1
  p_box_list = append(p_box_list, list(p_box))
  # paper_plots[[paste('box_logistic', top_bottom_q, sep = '_')]] = p_box
   
  #ggsave(file.path(plot_out_path, paste0(paste('logistic_box', top_bottom_q, this_top_quantile, metric, sep = '_'), '.png')), p_box, height = 6, width = 4)
}
box_combined = ggarrange(plotlist = p_box_list)
print(box_combined)
# paper_plots[['box_logistic']] = box_combined
```
```{r}
delta_df_logistic %>%
     filter(top_quantile == this_top_quantile) %>%
    filter(fdr == 0.05 & .metric == 'pr_auc') %>%
           filter(!! method_gene_list_filter) %>%
  group_by(rare_burden, outlier_mode_name) %>% 
  summarise(mean = mean(ratio_delta_auprc)*100, median = median(ratio_delta_auprc)*100) %>% 
  arrange(mean)
```


```{r}
#manually only plot NS comparison (DeepRVAT vs pLOF for bottom, all other are significant)
non_significant_comparisons_log = list(c("baseline-plof", "deeprvat-avg" ))
n = 0
p_box_list = list()
for (top_bottom_q in c('bottomq', 'topq')){

  plot_df = delta_df_logistic %>%
     filter(extreme_direction == top_bottom_q & top_quantile == this_top_quantile) %>%
         filter(fdr == 0.05 & .metric == 'pr_auc') %>%
           filter(!! method_gene_list_filter)
  
  y_min = min(plot_df[[metric]])
  y_max = quantile(plot_df[[metric]], 0.9999) #excluding extreme oultiers
  y_max = max(plot_df[[metric]][plot_df[[metric]] < y_max])
  pos_sig = seq(from = y_max, by = 0.03, length.out = n_items)
  pos_sig_2 = seq(from = y_max + 0.01 , by =  0.03, length.out = length(my_comparisons) - n_items)
  
  p_box = ggplot(#data=plot_df[plot_df[[metric]] <= y_max,],
                 data= plot_df,
             aes_string(y = metric, x = 'rare_burden', fill = 'rare_burden')) +
            geom_boxplot() +
            geom_hline(yintercept = 0, color = 'grey60') +
            facet_wrap(vars(outlier_mode_name), nrow = 1) +
            labs( y= expression(paste("Relative \u2206AUPRC")), 
                  fill = 'Rare variant \nscoring method', x = '') + # 'Relative  \u2206  R2'
            scale_x_discrete(limits = method_order,
                               labels = method_names) +
            scale_fill_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + 
            scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + 
            theme_cowplot() +
            scale_y_continuous(labels = scales::percent) +
           theme(axis.text.x = element_text(angle = 44,  hjust =1),
                         strip.background = facet_background)
  if(top_bottom_q == 'bottomq'){
      p_box = p_box + geom_signif(data = plot_df, color = 'grey30',
                        comparisons = non_significant_comparisons_log, 
                        map_signif_level=c("***"=0.001, "**"=0.01, "*"=0.05, "ns" = 0.51), 
                        test = "wilcox.test",
                        test.args = c(alternative = 'less'),
                        # y_position = c(pos_sig, pos_sig_2), 
                        # size = 0.2, vjust = 0.5,
                        tip_length = 0.01) 
  }
  if (n > 0){
    print('removing title')
    p_box = p_box + theme(axis.title.y = element_blank())
  }
  n = n + 1
  p_box_list = append(p_box_list, list(p_box))
  # paper_plots[[paste('box_logistic', top_bottom_q, sep = '_')]] = p_box
   
  #ggsave(file.path(plot_out_path, paste0(paste('logistic_box', top_bottom_q, this_top_quantile, metric, sep = '_'), '.png')), p_box, height = 6, width = 4)
}
box_combined = ggarrange(plotlist = p_box_list)
print(box_combined)
paper_plots[['box_logistic']] = box_combined
```

```{r}
this_extreme_quantile = '1%'
outlier_modes_to_keep = c('Top', 'Bottom')

p <- ggplot(data = ranked_df_combined_avg %>% 
  filter(extreme_quantile_str == this_extreme_quantile & outlier_mode %in% outlier_modes_to_keep)
  , aes(x = max_rank, y = sum, color = `Rare estimator`)) +
  # geom_point(size = 0.2) +
  geom_line() +
  facet_wrap(vars(outlier_mode_name),
             scales = 'free')     +
  scale_color_manual(values = color_mapping, guide = 'none') +
  geom_abline(slope = 1, intercept = 0, linetype = 'dotted', color = 'grey') +
  annotate("segment", x = -Inf, xend = Inf, y = -Inf, yend = -Inf) +
  annotate("segment", x = -Inf, xend = -Inf, y = -Inf, yend = Inf) +
  theme_cowplot() +
  theme(
        # strip.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = facet_background) +
  labs(y = 'Samples with \n extreme phenotype', x = TeX(r'(Rank ($\Delta_{common - rare~PRS})$)'))

p
paper_plots[['extreme_prediction']] = p
# Add x-axis and y-axis labels to the first plot in the gri

#ggsave(file.path(plot_out_path, paste0(paste('delta_outlier_in_extreme_enrichment_all_traits',  sub('%', '',this_extreme_quantile), sep = '_'), '.png')), p,  height = 3.5, width = 6)
```


```{r}
plot_data = enrich_score_avg_df %>% filter(min_z_score <= max_z_score)
p = ggplot(data = plot_data,
           aes(x = min_z_score, y = enrichment, color = model, linetype = prs_quantile)) +
    geom_line(size = 0.3) +
    theme_cowplot() +
    scale_color_manual(values = color_mapping, limits = names(color_mapping), labels = method_names[names(color_mapping)], breaks = names(color_mapping) ) + #, guide = NULL) + #limits = c(baseline_methods, 'deeprvat-avg') + 
    labs(x = 'Phenotype outlier (z-score)', y = 'Enrichment vs baseline', color = '', linetype = '') +
  theme(legend.position = c(0.05, 0.7),
        legend.text = element_text(size = 10),
        legend.spacing.y = unit(0.05, "cm"))

p
paper_plots[['extreme_enrichment']] = p
```

```{r}
rare_burdens = delta_df %>% distinct(rare_burden) %>% pull
p_legend = ggplot(data=delta_df %>% 
             filter(fdr == 0.05) %>%
               filter(rare_burden != 'missense') %>%
               filter(gene_list == 'deeprvat_discoveries' & .metric == 'rsq'), 
           aes(x=phenotype, y=delta_r2, fill = rare_burden)) +
          geom_bar(stat="identity", position=position_dodge()) + 
          scale_fill_manual(values = color_mapping[rare_burdens], limits = rare_burdens, labels = method_names[rare_burdens], breaks = rare_burdens) + 
          theme_cowplot() +
          scale_x_discrete(labels = phenotype_dict) +
          theme_cowplot() +
          theme(axis.text.x = element_text(angle = 44,  hjust =1),
                legend.position="bottom",
                legend.title = element_blank(),
                legend.box.margin = margin(0,0,0,1)) #+
          # guides(fill = guide_legend(nrow = 1))
print(p_legend)
g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

mylegend<-g_legend(p_legend)
mylegend
paper_plots[['legend']] = mylegend
```


## Arrange main figure
```{r}


(1/6) * 2
lay <- rbind(c(NA,NA,NA,NA,NA,NA,NA,NA,NA),
             c(1,1,1,2,2,2,2,2,2),
             c(NA,3,3,3,3,NA, 5, 5,5),
             c(4,4,4,4,4,NA,5, 5,5))
gt = grid.arrange(paper_plots[['linear_boxplot']],
             paper_plots[['box_logistic']],
            paper_plots[['legend']],
             paper_plots[['extreme_prediction']],
             paper_plots[['extreme_enrichment']],
          nrow = 4, ncol = 9,
          layout_matrix = lay, heights = c(0.1,1.3,0.1,1),
          widths = c(1,1,1,1,1,0.5,1,1,1))
p = as_ggplot(gt) +                                # transform to a ggplot
  draw_plot_label(label = c("a", "b", "c", "d"), size = 25,
                  x = c(0.00, 3/9, 0.05, 5.5/9), y = c(1, 1,0.47, 0.47)) # Add labels

ggsave(file.path(plot_out_path, 'figure_4.png'), p, width = 24, height = 20, unit = 'cm')

```


```{r}
plot_width = 36
# generated by figure_4_hold_out_phenotype_prediction.Rmd
p_rel_logistic = readRDS('p_rel_logistic.Rds')
p_rel_linear = readRDS( 'p_rel_linear.Rds')

p_rel_linear = p_rel_linear + theme(plot.margin = margin(0,0,0,20))
p_rel_logistic = p_rel_logistic + theme(plot.margin = margin(0,0,0,20))

(1/6) * 2
lay <- rbind(c(NA,NA,NA,NA,NA,NA,NA,NA,NA,NA),
             c(1,1,1,2,2,2,2,2,2,6),
             c(NA,3,3,3,3,NA, 5, 5,5,7),
             c(4,4,4,4,4,NA,5, 5,5, 7))
gt = grid.arrange(paper_plots[['linear_boxplot']],
             paper_plots[['box_logistic']],
            paper_plots[['legend']],
             paper_plots[['extreme_prediction']],
             paper_plots[['extreme_enrichment']],
            p_rel_linear, p_rel_logistic,
          nrow = 4, ncol = 10,
          layout_matrix = lay, heights = c(0.1,1.3,0.3,1),
          widths = c(1,1,1,1,1,0.2,1,1,1,3))
p = as_ggplot(gt) +                                # transform to a ggplot
  draw_plot_label(label = c("a", "b", "c", "d", 'e', 'f'), size = plot_width/default_plot_width * label_default_size,
                  x = c(0, 2.5/10, 0, 4.5/10, 7.3/10, 7.3/10), y = c(1, 1,0.41, 0.5, 1, 0.5)) # Add labels
p
ggsave(file.path(plot_out_path, 'figure_4_v2.png'), p, width = plot_width, height = 24, unit = 'cm')
```




# Supplements

## Gene list comparison
```{r}

p_box_lin_gene_list = ggplot(data=delta_df %>%
             filter(fdr == 0.05 & .metric == 'rsq'),
           aes(y=ratio_r2, x = rare_burden, fill = gene_list)) +
          geom_boxplot() +
          # rare_method_fill +
          labs( y= expression(paste("Relative \u2206", R^2)), fill = 'Gene list', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = c(head(baseline_methods, n_items), 'deeprvat-avg', tail(baseline_methods, -n_items)),
                             labels = method_names) +
          scale_fill_brewer(palette = 'Pastel1', limits = names(gene_list_renamer), labels = gene_list_renamer) +   
          theme_cowplot() +
          remove_y_expand +
          theme(axis.text.x = element_text(angle = 45,  hjust =1),
                strip.background = facet_background)
print(p_box_lin_gene_list)

```

```{r}
delta_df_logistic %>% filter(is.na(metric))
plot_df = delta_df_logistic %>%
   filter(top_quantile == this_top_quantile) %>%
       filter(fdr == 0.05 & .metric == 'pr_auc')

p_box_log_gene_list = ggplot(#data=plot_df[plot_df[[metric]] <= y_max,],
               data= plot_df,
           aes_string(y = metric, x = 'rare_burden', fill = 'gene_list')) +
          geom_boxplot() +
          geom_hline(yintercept = 0, color = 'grey60') +
          facet_wrap(vars(outlier_mode_name), nrow = 1) +
          labs( y= expression(paste("Relative \u2206AUPRC")), 
                fill = 'Gene list', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = c(head(baseline_methods, n_items), 'deeprvat-avg', tail(baseline_methods, -n_items)),
                           labels = method_names) +
          scale_fill_brewer(palette = 'Pastel1', limits = names(gene_list_renamer), labels = gene_list_renamer) +           # scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + #limits = c(baseline_methods, 'deeprvat-avg')
          theme_cowplot() +
          scale_y_continuous(labels = scales::percent) +
         theme(axis.text.x = element_text(angle = 44,  hjust =1),
               strip.background = facet_background)
print(p_box_log_gene_list)

```


```{r}
plot_width = 18
p_box_gene_list = ggarrange(p_box_lin_gene_list , 
          p_box_log_gene_list,   common.legend = TRUE,
          widths = c(0.4, 0.6),
          labels = 'auto',
          font.label = list(size = plot_width/default_plot_width * label_default_size),
          legend="bottom") 

ggsave(file.path(plot_out_path, 'figure_4_supp_1.png'), width = plot_width, height = 10, unit = 'cm',
       dpi = 600)
ggsave(file.path(plot_out_path, 'figure_4_supp_1.pdf'), width = plot_width, height = 10, unit = 'cm',
       dpi = 600)

```

## relative improvement enrichment (for fig. 4e)
```{r}
methods =  avg_rel_improvement_df %>% distinct(model) %>% filter(!grepl(paste(burdens_to_exclude, collapse = '|'), model)) %>% pull(model)
methods = 
p = ggplot(data = avg_rel_improvement_df %>% filter(min_z_score <= max_z_score),
         aes(x = min_z_score, y = rel_improvment_common, color = model)) +
  # geom_point(size = 0.2) +
  geom_line(size = 0.3) +
  facet_wrap(vars(prs_quantile), scales = 'free', ncol = 2) +
  theme_cowplot() +
  scale_color_manual(values = color_mapping[methods], limits = methods,
                     labels = method_names[methods]) +
  labs(x = 'Phenotype outlier (z-score)', y = 'Relative enrichment vs baseline', color = '', linetype = '') +
  scale_y_continuous(labels = scales::percent) +
  theme(strip.background = facet_background)


print(p)

ggsave(file.path(plot_out_path, 'figure_4_supp_2.png'), width = 18, height = 12, unit = 'cm',
       dpi = 600)
ggsave(file.path(plot_out_path, 'figure_4_supp_2.pdf'), width = 18, height = 12, unit = 'cm',
       dpi = 600)
```


















