---
title: "DeepRVAT - RVAT figures"
output:
  pdf_document: default
  html_notebook: default
---

Load libraries

```{r}

library(ggplot2)
library(arrow)
library(dplyr)
library(stringr)
library(tidyr)
library(ggpubr)
library(cowplot)
library(latex2exp)
library(tibble)

source('rvat_figure_utils.R')
pheno_choice = 'all_phenotypes'

```

# Define experiment directories
```{r}
## define directories
monti_res_binary_dir = 'monti_binary'
staar_res_binary_dir = 'staar_binary'
monti_quant_dir = 'monti'
staar_quant_dir = 'staar'
results_dir = 'paper_experiment' #deeprvat quantitative
results_deeprvat_binary_dir = 'deeprvat_binary'
timing_data_dir = '../timing/' #timing.Rmd has to be run before executing this markdown
```





```{r}
phenotypes = read_parquet('phenotypes.parquet')
binary_trait_counts <- colSums(na.omit(phenotypes %>% select(all_of(binary_phenotypes))))

colSums(!is.na(phenotypes %>% select(all_of(new_quant_phenotypes))))

tibble(Column = gsub('_', ' ', gsub('Jurgens_', '', names(binary_trait_counts))), count = binary_trait_counts)

```


# Load results

## DeepRVAT
```{r}

all_deeprvat_res = loadDeepRVATResults(results_dir = results_dir, phenotypes = quant_phenotypes, phenotype_renamer = phenotype_renamer)
counts_deeprvat_quant = all_deeprvat_res[['counts']]
significant_deeprvat_quant = all_deeprvat_res[['significant']]
results_deeprvat_quant = all_deeprvat_res[['results']]
pheno_names_to_keep = all_deeprvat_res[['pheno_names_to_keep']]
```



## Load Monti et al. results

```{r}
counts_monti_quant = readRDS(file.path(monti_quant_dir, paste0("monti_counts", monti_staar_name_dict[pheno_choice], ".Rds"))) %>%
  filter(Trait != 'WHR') %>%
  mutate(is_single_trait = ifelse(Trait == 'All traits', 'False', 'True')) %>%
  rename(n = discoveries) %>%
  mutate(Method = 'Monti et al.') %>%
  mutate(Trait = ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait)) %>%
  filter(Trait %in% pheno_names_to_keep)

monti_res_quant = readRDS(file.path(monti_quant_dir, paste0("monti_results", monti_staar_name_dict[pheno_choice], ".Rds")))  %>% 
  filter(Trait != 'WHR') %>%
  mutate(Method = 'Monti et al.') %>%
  mutate(Trait = ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait)) %>% 
  filter(Trait %in% pheno_names_to_keep)
monti_res_quant %>% distinct(Trait)


assertthat::assert_that(setequal(counts_deeprvat_quant %>% distinct(Trait) %>% pull(Trait), counts_monti_quant %>% distinct(Trait) %>% pull(Trait)))
assertthat::assert_that(setequal(results_deeprvat_quant %>% distinct(Trait) %>% pull(Trait), monti_res_quant %>% distinct(Trait) %>% pull(Trait)))

all_traits_count = sum(counts_monti_quant %>% filter(is_single_trait == 'True') %>% pull(n))
counts_monti_quant = counts_monti_quant %>% mutate(n = ifelse(Trait == 'All traits',all_traits_count, n))
```

## Load STAAR results


```{r}
counts_staar_quant = readRDS(file.path(staar_quant_dir, paste0("sig_counts_staar", monti_staar_name_dict[pheno_choice], ".Rds"))) %>% 
  filter(Trait != 'WHR') %>%
  mutate(Method = 'STAAR')  %>%
  rename(n = discoveries) %>%
  mutate(is_single_trait = ifelse(Trait == 'All traits', 'False', 'True')) %>%
  mutate(Trait = ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait)) %>%
  filter(Trait %in% pheno_names_to_keep)

all_traits_count = sum(counts_staar_quant %>% filter(is_single_trait == 'True') %>% pull(n))
counts_staar_quant = counts_staar_quant %>% mutate(n = ifelse(Trait == 'All traits',all_traits_count, n))


staar_res_quant = readRDS(file.path(staar_quant_dir, paste0("staar_corrected", monti_staar_name_dict[pheno_choice], ".Rds")))  %>% 
  filter(Trait != 'WHR') %>%
  mutate(Method = 'STAAR') %>%
  mutate(Trait = ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait)) %>% 
  filter(Trait %in% pheno_names_to_keep)  %>%
  drop_na(pval) %>%
  mutate(pval = as.numeric(pval))



assertthat::assert_that(setequal(counts_deeprvat_quant %>% distinct(Trait) %>% pull(Trait), counts_staar_quant %>% distinct(Trait) %>% pull(Trait)))
assertthat::assert_that(setequal(results_deeprvat_quant %>% distinct(Trait) %>% pull(Trait), staar_res_quant %>% distinct(Trait) %>% pull(Trait)))
```



## Combine quantiative counts results

```{r} 
all_counts_quant =  rbind(counts_deeprvat_quant, counts_monti_quant, counts_staar_quant) %>%
  mutate(Method = factor(Method, levels=methods)) %>%
  mutate(Trait = as.character(Trait)) %>%
mutate(Trait = recode(Trait, !!!phenotype_renamer))  %>%  
  mutate(pheno_group = case_when(
    Trait %in% names(old_quant_phenotypes) ~ 'old',
    Trait %in% names(new_quant_phenotypes) ~ 'new',
    TRUE ~ 'all'
  )) 

all_counts_quant %>% distinct(Trait, pheno_group)
all_counts_quant %>% filter(Trait == "All traits")
all_counts_quant %>% filter(Trait != "All traits")
levels(all_counts_quant$Method)
all_counts_quant
```

```{r}
all_counts_quant %>% distinct(pheno_group)
all_counts_quant %>% filter(pheno_group == 'Training Trait') %>% 
  filter(Method != 'STAAR')
```

## group quantiative counts
```{r}

flattened_quant_trait_grouping <- stack(quant_trait_grouping) %>%
  mutate(ind = gsub('\\d+', '', ind))
quant_trait_grouping = flattened_quant_trait_grouping [['ind']]
names(quant_trait_grouping) = flattened_quant_trait_grouping[['values']]
quant_trait_grouping

setdiff(all_counts_quant %>% distinct(Trait) %>% pull(Trait), names(quant_trait_grouping))
all_counts_quant = all_counts_quant %>% mutate(trait_group = quant_trait_grouping[as.character(Trait)]) %>%
  mutate(pheno_group = phenotype_old_new_dict[as.character(Trait)]) %>%
  mutate(pheno_group =  factor(pheno_group, levels = c('Training Trait', 'New Trait')))

grouped_counts_by_pheno_group = all_counts_quant %>% filter(is_single_trait == 'True') %>%
  group_by(Method, trait_group, pheno_group) %>%
  summarise(n = sum(n)) %>%
  ungroup()
grouped_counts_all_pheno_group = all_counts_quant %>% filter(is_single_trait == 'True') %>%
  group_by(Method, pheno_group) %>%
  summarise(n = sum(n)) %>%
  ungroup() %>%
  mutate(trait_group = 'All Traits')
grouped_counts = rbind(grouped_counts_by_pheno_group, grouped_counts_all_pheno_group)  
grouped_counts %>% filter(pheno_group == 'All Traits')

trait_counts = all_counts_quant %>% filter(is_single_trait == 'True') %>%
  select(pheno_group, Trait, trait_group) %>%
  distinct() %>%
  group_by(trait_group, pheno_group) %>%
  summarise(n_phenotypes = n()) %>%
  mutate(trait_with_count = paste0(trait_group, ' (',n_phenotypes, ')' ))

trait_counts = rbind(trait_counts, all_counts_quant %>% filter(is_single_trait == 'True') %>%
  select(pheno_group, Trait) %>%
  distinct() %>%
  group_by(pheno_group) %>%
  summarise(n_phenotypes = n()) %>% 
  mutate(trait_group = 'All Traits')) %>%
  mutate(trait_with_count = paste0(trait_group, ' (',n_phenotypes, ')' ))

trait_counts

grouped_counts = grouped_counts %>% left_join(trait_counts)
grouped_counts = grouped_counts %>% mutate(is_single_trait = ifelse(trait_group  == 'All Traits', FALSE, TRUE)) 
```

# Figure 3 plots

```{r}
plot_list = list()

```

```{r}
trait_title_dict = c('Training Trait' = 'Traits used in gene impairment module training', 'New Trait' = 'Traits not used in gene impairment module training')
for (t_group in c('Training Trait', 'New Trait')){
  p = ggplot(
        grouped_counts %>% filter(pheno_group == t_group & Method %in% main_methods),
        aes(
          x=trait_with_count,
          y=n,
          fill=Method,
        )
    ) + 
    geom_col(position='dodge', width=0.9, color='darkgray', linewidth=0.1) +
    labs(y = "Significant gene-trait associations", title = trait_title_dict[[t_group]]) +
    theme_classic() +
    # facet_grid(.~is_single_trait, scales='free', space = 'free') + #,  space = "free_x") +
    ggforce::facet_row(vars(is_single_trait), scales = 'free', space = 'free') +
    # theme_cowplot() +
    base_theme +
      theme(
        axis.text.x=element_text(angle=45, vjust=1, hjust=1),
        plot.title = element_text(hjust = 0.5, size = title_font_size),
        axis.title.x=element_blank(),
        legend.key.size=unit(0.5, "line"),
        legend.position = "none",
        panel.spacing=unit(0.75, "cm"),
         strip.background = element_blank(),
        strip.text = element_blank(),
        plot.margin = margin(1, 1, 1, 1, "cm"),
    ) +
    scale_fill_manual(values = colors) +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) 
  print(p)
  plot_list[[paste('quant_discoveries', gsub(' ', '_', t_group), sep = "_")]] = p
}
grouped_counts %>% filter(trait_group == 'All Traits')
(155 - 69)/69
```




# Replication training traits
Here we only check the replication on the training phenotypes

```{r}
pheno_suffix = '_training_phenotypes'

replication_file_staar = file.path(staar_quant_dir, paste0('replication_staar', pheno_suffix, '.Rds'))
replication_file_monti = file.path(monti_quant_dir, paste0('replication_monti', pheno_suffix, '.Rds'))
replication_file_deeprvat = file.path(results_dir, "replication" ,paste0("replication", pheno_suffix, ".parquet"))
replication_data = readReplicationData(replication_file_staar, replication_file_monti, replication_file_deeprvat)
replication_plot = makeReplicationPlot(replication_data %>% filter(Method %in% main_methods), max_rank = 1000)
replication_plot

plot_list[['replication_training']] = replication_plot
```



# binary phenotypes
```{r}
phenotypes_df = read_parquet('phenotypes.parquet') %>%
   mutate_all(as.numeric) %>%
  mutate_all(~replace(., is.na(.), 0))
case_counts = colSums(phenotypes_df[binary_phenotypes])
case_counts = tibble('Trait' = names(case_counts), 'n_cases' = case_counts)
case_counts = case_counts %>%  mutate(Trait = str_replace_all(Trait, "_", " "))

```

```{r}
all_deeprvat_binary_res = loadDeepRVATResults(results_dir = results_deeprvat_binary_dir, 
                                       phenotypes =  gsub(' ', '_', binary_phenotypes), 
                                       phenotype_renamer = c())
counts_deeprvat_binary = all_deeprvat_binary_res[['counts']]
significant_deeprvat_binary = all_deeprvat_binary_res[['significant']]
results_deeprvat_binary = all_deeprvat_binary_res[['results']]
pheno_names_to_keep = all_deeprvat_binary_res[['pheno_names_to_keep']]

#add zero counts for traits that are zero for all methods so we don't loose them
zero_count_traits = setdiff(unique(results_deeprvat_binary[['Trait']]), unique(counts_deeprvat_binary[['Trait']]))
for (trait in zero_count_traits){
  print(paste('adding zero counts for Trait', trait))
  counts_deeprvat_binary = rbind(counts_deeprvat_binary, 
                           counts_deeprvat_binary %>% filter(Trait == 'All traits') %>% mutate(Trait = trait) %>%
    mutate(n = 0))
}

counts_deeprvat_binary[is.na(counts_deeprvat_binary$is_single_trait), "is_single_trait"] <- "True"

lambda_deeprvat_binary = compute_lambda(results_deeprvat_binary %>% group_by(Trait, Method)) %>% ungroup()
counts_deeprvat_binary_with_qc = counts_deeprvat_binary %>% left_join(lambda_deeprvat_binary) %>% 
  left_join(case_counts) %>%
  select(Trait, Method, n,lambda_gc, n_cases) %>% 
  arrange(desc(lambda_gc))
```

```{r}
counts_deeprvat_binary_with_qc
plot_counts = counts_deeprvat_binary_with_qc %>% 
  filter(grepl('Jurgens', Trait)) %>%
  mutate(Trait = gsub('Jurgens ', '', Trait))

```

```{r}
binary_trait_grouping_tibble
```


```{r}

plot_counts_grouped = plot_counts %>% 
  left_join(binary_trait_grouping_tibble) %>% 
  group_by(binary_trait_grouping, Method) %>% 
  summarise(n = sum(n), n_conditions = n()) %>%
  ungroup()
n_binary_traits = sum(plot_counts_grouped %>% select(binary_trait_grouping, n_conditions) %>% distinct() %>% pull(n_conditions))

plot_counts_grouped_all_traits = plot_counts_grouped %>% group_by(Method) %>% 
  summarise(n = sum(n)) %>% 
  ungroup() %>%
  mutate(n_conditions = n_binary_traits) %>%
  mutate(binary_trait_grouping = 'All Traits')

plot_counts_grouped = rbind(plot_counts_grouped, plot_counts_grouped_all_traits) %>% 
  mutate(trait_with_count = paste0(binary_trait_grouping, ' (', n_conditions,  ')')) %>%
  filter(Method %in% binary_methods) %>%
  mutate(Method = factor(Method, levels = binary_methods)) %>%
  mutate(is_single_trait = ifelse(binary_trait_grouping == 'All Traits', FALSE, TRUE))
```


# Calibration on new phenotypes (Binary and quantitative)

## Load Monti and STAAR pvals for binary traits

```{r}
monti_res_binary = readRDS(file.path(monti_res_binary_dir, 'monti_results.Rds')) %>%
  mutate(Trait = str_replace_all(Trait, "_", " ")) %>%
  mutate(Method = 'Monti et al.')

counts_monti_binary =  readRDS(file.path(monti_res_binary_dir, "monti_counts.Rds")) %>% 
  mutate(Method = 'Monti et al.') %>%
  mutate(Trait = str_replace_all(gsub('Jurgens_', '', Trait), "_", " ")) %>%
  rename(n = discoveries) %>%
  filter(Trait != 'All traits')

staar_res_binary = readRDS(file.path(staar_res_binary_dir, "staar_corrected.Rds")) %>%
  mutate(Trait = str_replace_all(Trait, "_", " ")) %>%
  drop_na(pval) %>%
  mutate(pval = as.numeric(pval)) %>%
  mutate(Method = 'STAAR') 

counts_staar_binary = readRDS(file.path(staar_res_binary_dir, "sig_counts_staar.Rds")) %>% 
  mutate(Method = 'STAAR')  %>%
  rename(n = discoveries) %>%
  mutate(Trait = str_replace_all(gsub('Jurgens_', '', Trait), "_", " ")) %>%
  filter(Trait != 'All traits')
```


```{r}
pval_cols = c('Trait', 'gene', 'pval', 'Method')
all_binary_pvals = rbind(results_deeprvat_binary[pval_cols],
                         monti_res_binary[pval_cols], staar_res_binary[pval_cols])%>%
  mutate(trait_group = 'Binary')

new_phenotypes = c(all_binary_pvals %>% distinct(Trait) %>% pull(Trait),  names(new_quant_phenotypes))
all_pvals_quant = rbind(results_deeprvat_quant %>% 
                            select(all_of(pval_cols)),
                           monti_res_quant %>% select(all_of(pval_cols)), 
                           staar_res_quant %>% select(all_of(pval_cols))
                           ) %>%
  mutate(trait_group = 'Quantitative')

write_parquet(all_pvals_quant, 'all_pvals_quant.parquet')
all_pvals = rbind(all_binary_pvals, all_pvals_quant)

all_lambda = compute_lambda(all_pvals %>% group_by(Trait, Method, trait_group)) %>% ungroup() %>%
  mutate(Method = factor(Method, levels = methods)) %>%
  mutate(Trait = gsub('Jurgens ', '', Trait))
all_lambda %>% distinct(Trait)


```

```{r}
assertthat::assert_that(setequal(plot_counts %>% distinct(Trait) %>% pull(Trait), counts_staar_binary %>% distinct(Trait) %>% pull(Trait)))
assertthat::assert_that(setequal(plot_counts %>% distinct(Trait) %>% pull(Trait), counts_monti_binary %>% distinct(Trait) %>% pull(Trait)))
counts_staar_binary
binary_counts_with_qc = rbind(counts_monti_binary, counts_staar_binary) %>% left_join(all_lambda %>% select(Trait, Method, lambda_gc)) %>%
  left_join(plot_counts %>% select(Trait, n_cases)) %>%
  rbind(plot_counts)
saveRDS(binary_counts_with_qc, 'binary_counts_with_qc.Rds') # used for plotting supp. figures
```


# Generate all plots

```{r}
trait_title_dict = c('Training Trait' = 'Traits used in gene impairment module training', 'New Trait' = 'Traits not used in gene impairment module training')
for (t_group in c('Training Trait', 'New Trait')){
  p = ggplot(
        grouped_counts %>% filter(pheno_group == t_group & Method %in% main_methods),
        aes(
          x=trait_with_count,
          y=n,
          fill=Method,
        )
    ) + 
    geom_col(position='dodge', width=0.9, color='darkgray', linewidth=0.1) +
    labs(y = "Significant gene-trait associations", title = trait_title_dict[[t_group]]) +
    theme_classic() +
    # facet_grid(.~is_single_trait, scales='free', space = 'free') + #,  space = "free_x") +
    ggforce::facet_row(vars(is_single_trait), scales = 'free', space = 'free') +
    base_theme_wo_strip +
      theme(
        axis.text.x=element_text(angle=45, vjust=1, hjust=1),
        plot.title = element_text(hjust = 0.5, size = title_font_size),
        axis.title.x=element_blank(),
        legend.key.size=unit(0.5, "line"),
        legend.text = element_text(size = 12),
        panel.spacing=unit(0.75, "cm"),
         strip.background = element_blank(),
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) 
  if (t_group == 'Training Trait'){
    p = p + 
      scale_fill_manual(values = colors,  guide = guide_legend(
      direction = "horizontal", nrow = 2
    )) + theme(legend.position = c(0.65, 0.8)) 
  }else{
    p = p + theme(legend.position = 'none') +
       scale_fill_manual(values = colors) 
  }
  
  print(p)
  plot_list[[paste('quant_discoveries', gsub(' ', '_', t_group), sep = "_")]] = p
}
grouped_counts %>% filter(trait_group == 'All Traits')
```


```{r}
replication_plot = makeReplicationPlot(replication_data %>% filter(Method %in% main_methods), max_rank = 1000)
replication_plot

plot_list[['replication_training']] = replication_plot 
```


```{r}
counts_grouped_traits = ggplot(
      plot_counts_grouped %>%
         filter(Method %in% c( 'Burden/SKAT combined', 'DeepRVAT')),
      aes(
        x=trait_with_count,
        y=n,
        fill=Method,
      )
  ) + 
  geom_col(position='dodge', width=0.9, color='darkgray', linewidth=0.1) +
  labs(y = "Significant gene-trait associations", title = 'Binary traits') +
  theme_classic() +
  # facet_grid(.~is_single_trait, scales='free', space = 'free') + #,  space = "free_x") +
  ggforce::facet_row(vars(is_single_trait), scales = 'free', space = 'free') +
  base_theme_wo_strip +
    theme(
      axis.text.x=element_text(angle=45, vjust=1, hjust=1),
      plot.title = element_text(hjust = 0.5, size = title_font_size),
      axis.title.x=element_blank(),
      legend.key.size=unit(0.5, "line"),
      legend.position = "none",
      panel.spacing=unit(0.75, "cm"),
       strip.background = element_blank(),
      strip.text = element_blank()
      # plot.margin = margin(1, 1, 1, 1, "cm"),
  ) +
  scale_fill_manual(values = colors) +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) 
counts_grouped_traits

binary_legend = ggplot(
      plot_counts_grouped %>%
         filter(Method %in% c( 'Burden/SKAT combined', 'DeepRVAT')),
      aes(
        x=trait_with_count,
        y=n,
        fill=Method,
      )
  ) + 
  geom_col(position='dodge', width=0.9, color='darkgray', linewidth=0.1) +
  labs(y = "Significant gene-trait associations", title = 'Binary traits') +
  theme_classic() +
  # facet_grid(.~is_single_trait, scales='free', space = 'free') + #,  space = "free_x") +
  ggforce::facet_row(vars(is_single_trait), scales = 'free', space = 'free') +
  base_theme_wo_strip +
    theme(
      axis.text.x=element_text(angle=45, vjust=1, hjust=1),
      plot.title = element_text(hjust = 0.5),
      axis.title.x=element_blank(),
      legend.key.size=unit(0.5, "line"),
      legend.position = "none",
      panel.spacing=unit(0.75, "cm"),
       strip.background = element_blank(),
      strip.text = element_blank()
      # plot.margin = margin(1, 1, 1, 1, "cm"),
  ) +
  scale_fill_manual(values = colors) +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) 
binary_legend
plot_list[['binary_counts']] = counts_grouped_traits
```


```{r}
main_methods = c('Burden/SKAT combined', "Monti et al.", "STAAR", 'DeepRVAT')
main_methods_short = c('B/S', 'M', 'ST', 'D')
main_methods_short = c('Burden/SKAT \n combined', "Monti et al.", "STAAR", 'DeepRVAT')
names(main_methods_short) = main_methods
p_lambda = ggplot(all_lambda %>% filter(Method %in% main_methods) %>% 
                  filter(Trait %in% new_phenotypes), 
                  aes(x = Method , y = lambda_gc, color = Method)) +
  facet_wrap(vars(trait_group)) +
  geom_jitter() +
  geom_boxplot() +
  # theme_cowplot() +
  theme_classic() +
  base_theme+
  theme(
      axis.text.x = element_text(angle = 45,  hjust =1, size = font_size)
      ) +
  scale_color_manual(values = colors) +
  scale_x_discrete(breaks = names(main_methods_short), labels = main_methods_short) +
  guides(color = FALSE) +
  labs(x = '', y = 'Lambda GC')
p_lambda
```

```{r}
df = all_pvals %>% filter(Trait %in% new_phenotypes) %>% filter(Method %in% main_methods)
df = df %>%
  group_by(Method, trait_group) %>%
  arrange(pval) %>%
  mutate(`-log10pval_expected` = -log10(1:n() / n())) %>%
  ungroup()

# Check if '-log10pval' column exists, if not, add it
if (!('-log10pval' %in% colnames(df))) {
  df <- df %>%
    mutate(`-log10pval` = -log10(pval))
}


```

```{r}
#sample data to make plotting faster
sample_data <- function(df) {
  df %>%
    group_by(Method) %>%
    mutate(sample_percent = ifelse(`-log10pval_expected` < 2, 0.10, 0.5)) %>%
    filter(runif(n()) < sample_percent)
}

# Sample the data --> Comment this in if you want to make the qqplot plotting faster
# df_sub <- df %>%
#   sample_data()
df_sub = df
nrow(df)
#show all values > max_val as triangles
max_val_binary = 30
max_val_quantitative = 50

df_sub <- df_sub %>%
  mutate(
    shape = case_when(
      trait_group == 'Binary' &  `-log10pval` > max_val_binary ~ 'triangle',
      trait_group == 'Binary' &  `-log10pval` <= max_val_binary ~ 'circle',
      trait_group == 'Quantitative' &  `-log10pval` > max_val_quantitative ~ 'triangle',
      trait_group == 'Quantitative' &  `-log10pval` <= max_val_quantitative ~ 'circle',
      TRUE ~ NA_character_
    ),
     `-log10pval` = case_when(
      (trait_group == 'Binary' &  `-log10pval` > max_val_binary) |
      (trait_group == 'Quantitative' &  `-log10pval` > max_val_quantitative) ~ max_val_binary,
      TRUE ~  `-log10pval`
    )
  ) %>%
  mutate(alpha = ifelse(`-log10pval_expected` < 4, 'alpha_low', 'alpha_high'))
qq_p = ggplot(df_sub, aes(x = `-log10pval_expected`, y = `-log10pval`, color = Method, fill = Method, shape = shape, size = shape)) + #, alpha = `-log10pval_expected`)) +
    labs(x = TeX("$Expected -log_{10}(P)$"), y = TeX("$Observed -log_{10}(P)$")) +
    facet_wrap(vars(trait_group), scales = 'free') +
    geom_abline(intercept = 0, slope = 1, color = 'darkgrey') +
    geom_point(alpha = 0.6) +
    ylim(0, 30) +
    theme_classic() +
    base_theme_wo_margin +
    theme(legend.position = 'none',#,
          ) +
    scale_color_manual(values = colors) +
    scale_fill_manual(values = colors) +
    scale_shape_manual(values = c('triangle' = 17,'circle' = 19)) +
    scale_size_manual(values = c('triangle' = 1.5,'circle' = 0.5)) 


```




## timing plot
```{r}
this_df_x_fold = readRDS(file.path(timing_data_dir, paste0('timing_data_x_fold_60_pheno.Rds'))) #generate using deeprvat-analysis/timing/timing.Rmd

plot_colors_deeprvat = c('Monti' = "aquamarine4", 'STAAR' = '#E6AB02','DeepRVAT (with training)' = '#7A2961', 'DeepRVAT (pre-trained)' = '#7A2961')
plot_rename = c('Monti' = "Monti", 'STAAR' = 'STAAR','DeepRVAT (with training)' = 'DeepRVAT\n(with training)', 'DeepRVAT (pre-trained)' = 'DeepRVAT\n(pre-trained)')
days = c(1,  4,  16, 32, 64, 128)
days_sec = days * (3600*24)

# p_pheno_x_fold = 
 p_pheno_x_fold = ggplot(this_df_x_fold %>% filter(mode == 'time_test' ) ,
       aes(x = Method, y = time, color = Method, fill = Method)) +
  geom_bar(stat="identity") +
  geom_line() +
  scale_color_manual(values = plot_colors_deeprvat, guide = 'none') + 
  scale_fill_manual(values = plot_colors_deeprvat, guide = 'none') + 
  geom_text(aes(label=time_x_val),vjust=-1, size = 3.5) + 
  labs(y = paste('Testing time for 60 phenotypes [d]'), x = '', color = '') +
  scale_y_continuous(breaks = days_sec, labels = paste0(days, 'd'), expand = expansion(mult = c(0,0.3)))+ #, limits = c(0, max(days_sec))) +
  theme_classic() +
  base_theme_wo_margin +
  scale_x_discrete(breaks = names(plot_rename), labels = plot_rename) +
  theme(
        axis.text.x = element_text(angle = 45,  hjust =1, size = font_size)
        )

p_pheno_x_fold

```




```{r}
label_size = 25
label_specs = list(size = label_size, color = "black", face = "bold", family = NULL)
p = ggarrange(
  ggarrange(plot_list[['quant_discoveries_Training_Trait']], plot_list[['replication_training']] + theme(plot.margin = margin(1, 1, 2.5, 1, "cm")), align = 'h', widths = c(0.6, 0.4), labels = c('a', 'b'), font.label = label_specs,
            nrow = 1),
ggarrange(NA, NA, widths = c(0.235, 0.8), labels = c('c','d'),font.label = label_specs),
ggarrange(p_pheno_x_fold, p_lambda, qq_p + theme(plot.margin = margin(0.1, 0.25, 1.6, 0.25, "cm")), 
          widths = c(0.25,  0.3, 0.55), align = 'h', nrow = 1) , #labels = c('c', NA, 'd', NA),font.label = label_specs),
ggarrange(plot_list[['quant_discoveries_New_Trait']], plot_list[['binary_counts']], align = 'h', labels = c('e', 'f'), font.label = label_specs), 
nrow = 4, heights = c(0.95, 0.1, 0.75, 0.95)
)

width = 12
# ggsave('figure3.jpg', p, width = width , height = 1.2*width)
# ggsave('figure3.png', p, width = width , height = 1.2*width)

```


