---
title: "DeepRVAT - Figure 2: Benchmarking on unrelated Caucasian individuals from the 200k WES release"
author: "Eva Holtkamp"
output:
  pdf_document: default
  html_notebook: default
---

```{r}

library(ggplot2)
library(arrow)
library(dplyr)
library(stringr)
library(tidyr)
library(ggpubr)
library(cowplot)
library(latex2exp)
library(tibble)


source(file.path('rvat_figure_utils.R'))
source(file.path('../phenotypes.R'))

```

# Define experiment directories

```{r}

monti_quant_dir = '../comparison_methods/monti/experiments/monti/'  #directory where Monti has been run on unrelated Caucasian individuals from the 200k WES cohort
staar_quant_dir = '../comparison_methods/staar/experiments/staar/' #directory where STAAR has been run on unrelated Caucasian individuals from the 200k WES cohort
deeprvat_dir = "./deeprvat_main_exp/" #directory where DeepRVAT has been run on unrelated Caucasian individuals from the 200k WES cohort

deeprvat_dir_syn = './deeprvat_synonymous/' #directory where the DeepRVAT model from deeprvat_dir has been applied for association testing on synonymous variants only (unrelated Caucasian individuals from the 200k WES cohort) 

timing_data_dir = '../timing/' #directory where the timing experiments have been run 

scatter_data_dir = "" # DeepRVAT scores extracted for the plotted gene 
```



# Load results

## DeepRVAT
```{r}

all_deeprvat_res = loadDeepRVATResults(results_dir = deeprvat_dir, phenotypes = quant_phenotypes, phenotype_renamer = phenotype_renamer)
counts_deeprvat_quant = all_deeprvat_res[['counts']]
significant_deeprvat_quant = all_deeprvat_res[['significant']]
results_deeprvat_quant = all_deeprvat_res[['results']]
pheno_names_to_keep = all_deeprvat_res[['pheno_names_to_keep']]
```



## Load Monti et al. results

```{r}
counts_monti_quant = readRDS(file.path(monti_quant_dir,"monti_counts_all_phenotypes.Rds")) %>%
  mutate(is_single_trait = ifelse(Trait == 'All traits', 'False', 'True')) %>%
  rename(n = discoveries) %>%
  mutate(Method = 'Monti et al.') %>%
  mutate(Trait = ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait)) %>%
  filter(Trait %in% pheno_names_to_keep)

monti_res_quant = readRDS(file.path(monti_quant_dir, "monti_results_all_phenotypes.Rds"))  %>% 
  mutate(Method = 'Monti et al.') %>%
  mutate(Trait = ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait)) %>% 
  filter(Trait %in% pheno_names_to_keep)
monti_res_quant %>% distinct(Trait)


assertthat::assert_that(setequal(counts_deeprvat_quant %>% distinct(Trait) %>% pull(Trait), counts_monti_quant %>% distinct(Trait) %>% pull(Trait)))
assertthat::assert_that(setequal(results_deeprvat_quant %>% distinct(Trait) %>% pull(Trait), monti_res_quant %>% distinct(Trait) %>% pull(Trait)))

all_traits_count = sum(counts_monti_quant %>% filter(is_single_trait == 'True') %>% pull(n))
counts_monti_quant = counts_monti_quant %>% mutate(n = ifelse(Trait == 'All traits',all_traits_count, n))
```

## Load STAAR results

```{r}
counts_staar_quant = readRDS(file.path(staar_quant_dir, "sig_counts_staar_all_phenotypes.Rds")) %>% 
  mutate(Method = 'STAAR')  %>%
  rename(n = discoveries) %>%
  mutate(is_single_trait = ifelse(Trait == 'All traits', 'False', 'True')) %>%
  mutate(Trait = ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait)) %>%
  filter(Trait %in% pheno_names_to_keep)

all_traits_count = sum(counts_staar_quant %>% filter(is_single_trait == 'True') %>% pull(n))
counts_staar_quant = counts_staar_quant %>% mutate(n = ifelse(Trait == 'All traits',all_traits_count, n))


staar_res_quant = readRDS(file.path(staar_quant_dir, "staar_corrected_all_phenotypes.Rds")) %>% 
  mutate(Method = 'STAAR') %>%
  mutate(Trait = ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait)) %>% 
  filter(Trait %in% pheno_names_to_keep)  %>%
  drop_na(pval) %>%
  mutate(pval = as.numeric(pval))



assertthat::assert_that(setequal(counts_deeprvat_quant %>% distinct(Trait) %>% pull(Trait), counts_staar_quant %>% distinct(Trait) %>% pull(Trait)))
assertthat::assert_that(setequal(results_deeprvat_quant %>% distinct(Trait) %>% pull(Trait), staar_res_quant %>% distinct(Trait) %>% pull(Trait)))
```



## Combine quantiative counts results

```{r} 
all_counts_quant =  rbind(counts_deeprvat_quant, counts_monti_quant, counts_staar_quant) %>%
  mutate(Method = factor(Method, levels=methods)) %>%
  mutate(Trait = as.character(Trait)) %>%
mutate(Trait = recode(Trait, !!!phenotype_renamer))  %>%  
  mutate(pheno_group = case_when(
    Trait %in% names(old_quant_phenotypes) ~ 'old',
    Trait %in% names(new_quant_phenotypes) ~ 'new',
    TRUE ~ 'all'
  )) 

all_counts_quant %>% distinct(Trait, pheno_group)
all_counts_quant %>% filter(Trait == "All traits")
all_counts_quant %>% filter(Trait != "All traits")
levels(all_counts_quant$Method)
all_counts_quant
```

```{r}
all_counts_quant %>% distinct(pheno_group)
all_counts_quant %>% filter(pheno_group == 'Training Trait') %>% 
  filter(Method != 'STAAR')
```

## group quantiative counts
```{r}

flattened_quant_trait_grouping <- stack(quant_trait_grouping) %>%
  mutate(ind = gsub('\\d+', '', ind))
quant_trait_grouping = flattened_quant_trait_grouping [['ind']]
names(quant_trait_grouping) = flattened_quant_trait_grouping[['values']]
quant_trait_grouping

setdiff(all_counts_quant %>% distinct(Trait) %>% pull(Trait), names(quant_trait_grouping))
all_counts_quant = all_counts_quant %>% mutate(trait_group = quant_trait_grouping[as.character(Trait)]) %>%
  mutate(pheno_group = phenotype_old_new_dict[as.character(Trait)]) %>%
  mutate(pheno_group =  factor(pheno_group, levels = c('Training Trait', 'New Trait')))

grouped_counts_by_pheno_group = all_counts_quant %>% filter(is_single_trait == 'True') %>%
  group_by(Method, trait_group, pheno_group) %>%
  summarise(n = sum(n)) %>%
  ungroup()
grouped_counts_all_pheno_group = all_counts_quant %>% filter(is_single_trait == 'True') %>%
  group_by(Method, pheno_group) %>%
  summarise(n = sum(n)) %>%
  ungroup() %>%
  mutate(trait_group = 'All Traits')
grouped_counts = rbind(grouped_counts_by_pheno_group, grouped_counts_all_pheno_group)  
grouped_counts %>% filter(pheno_group == 'All Traits')

trait_counts = all_counts_quant %>% filter(is_single_trait == 'True') %>%
  select(pheno_group, Trait, trait_group) %>%
  distinct() %>%
  group_by(trait_group, pheno_group) %>%
  summarise(n_phenotypes = n()) %>%
  mutate(trait_with_count = paste0(trait_group, ' (',n_phenotypes, ')' ))

trait_counts = rbind(trait_counts, all_counts_quant %>% filter(is_single_trait == 'True') %>%
  select(pheno_group, Trait) %>%
  distinct() %>%
  group_by(pheno_group) %>%
  summarise(n_phenotypes = n()) %>% 
  mutate(trait_group = 'All Traits')) %>%
  mutate(trait_with_count = paste0(trait_group, ' (',n_phenotypes, ')' ))

trait_counts

grouped_counts = grouped_counts %>% left_join(trait_counts)
grouped_counts = grouped_counts %>% mutate(is_single_trait = ifelse(trait_group  == 'All Traits', FALSE, TRUE)) 
```

```{r}
grouped_counts %>% filter(trait_group == "All Traits") %>% arrange(pheno_group)
```

```{r}
## relative improvement

(272 - 155)/155
(272 - 228	)/228	
(272 - 255	)/255	
```


# Figure 3 plots

```{r}
plot_list = list()
```

```{r}
trait_title_dict = c('Training Trait' = 'Traits used in gene impairment module training', 'New Trait' = 'Traits not used in gene impairment module training')
for (t_group in c('Training Trait', 'New Trait')){
  p = ggplot(
        grouped_counts %>% filter(pheno_group == t_group & Method %in% main_methods),
        aes(
          x=trait_with_count,
          y=n,
          fill=Method,
        )
    ) + 
    geom_col(position='dodge', width=0.9, color='darkgray', linewidth=0.1) +
    labs(y = "Significant gene-trait associations", title = trait_title_dict[[t_group]]) +
    theme_classic() +
    # facet_grid(.~is_single_trait, scales='free', space = 'free') + #,  space = "free_x") +
    ggforce::facet_row(vars(is_single_trait), scales = 'free', space = 'free') +
    # theme_cowplot() +
    base_theme +
      theme(
        axis.text.x=element_text(angle=45, vjust=1, hjust=1),
        plot.title = element_text(hjust = 0.5, size = title_font_size),
        axis.title.x=element_blank(),
        legend.key.size=unit(0.5, "line"),
        legend.position = "none",
        panel.spacing=unit(0.75, "cm"),
         strip.background = element_blank(),
        strip.text = element_blank(),
        plot.margin = margin(1, 1, 1, 1, "cm"),
    ) +
    scale_fill_manual(values = colors) +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) 
  print(p)
  plot_list[[paste('quant_discoveries', gsub(' ', '_', t_group), sep = "_")]] = p
}
grouped_counts %>% filter(trait_group == 'All Traits')
```




# Replication training traits
Here we only check the replication on the training phenotypes

```{r}
pheno_suffix = '_training_phenotypes'

replication_file_staar = file.path(staar_quant_dir, paste0('replication_staar', pheno_suffix, '.Rds'))
replication_file_monti = file.path(monti_quant_dir, paste0('replication_monti', pheno_suffix, '.Rds'))
replication_file_deeprvat = file.path(deeprvat_dir, "replication.parquet")
replication_data = readReplicationData(replication_file_staar, replication_file_monti, replication_file_deeprvat)
replication_plot = makeReplicationPlot(replication_data %>% filter(Method %in% main_methods), max_rank = 500)
replication_plot = replication_plot +  theme( 
        legend.position = c(0.6, 0.3),
        legend.title = element_text(size = font_size, family = font_family)) +
  guides(color = 'none')
replication_plot

plot_list[['replication_training']] = replication_plot
```


```{r}
pheno_suffix = '_new_phenotypes'

replication_file_staar = file.path(staar_quant_dir, paste0('replication_staar', pheno_suffix, '.Rds'))
replication_file_monti = file.path(monti_quant_dir, paste0('replication_monti', pheno_suffix, '.Rds'))
replication_file_deeprvat  = file.path(deeprvat_dir, paste0("replication", pheno_suffix, ".parquet"))

replication_data = readReplicationData(replication_file_staar, replication_file_monti, replication_file_deeprvat)
replication_plot = makeReplicationPlot(replication_data %>% filter(Method %in% main_methods), max_rank = 150)
replication_plot = replication_plot +  theme( 
        legend.position = c(0.6, 0.3),
        legend.title = element_text(size = font_size, family = font_family)) +
  guides(color = 'none')
replication_plot
plot_list[['replication_new']] = replication_plot
```







# Calibration on new phenotypes (Quantitative)



## load pvals for deeprvat applied to syn variants only

```{r}
all_deeprvat_res_syn = loadDeepRVATResults(results_dir = deeprvat_dir_syn, 
                                       phenotypes = quant_phenotypes, 
                                       phenotype_renamer = phenotype_renamer)

results_deeprvat_quant_syn = all_deeprvat_res_syn[['results']]
counts_deeprvat_quant_syn = all_deeprvat_res_syn[['counts']]
```



```{r}
pval_cols = c('Trait', 'gene', 'pval', 'Method')

all_pvals_quant = rbind(results_deeprvat_quant %>% 
                            select(all_of(pval_cols)),
                           monti_res_quant %>% select(all_of(pval_cols)), 
                           staar_res_quant %>% select(all_of(pval_cols)), 
                        results_deeprvat_quant_syn %>% filter(Method == 'DeepRVAT')%>%
                            select(all_of(pval_cols)) %>% mutate(Method = 'DeepRVAT-Synonymous'
                                                                )) %>%
  mutate(trait_group = 'Quantitative')

all_pvals =  all_pvals_quant 

all_pvals %>% distinct(Method)
```



```{r}
df = all_pvals %>% filter(Trait %in% names(old_quant_phenotypes)) %>% filter(Method %in% c(main_methods,'DeepRVAT-Synonymous' ))
df = df %>%
  group_by(Method, trait_group) %>%
  arrange(pval) %>%
  mutate(`-log10pval_expected` = -log10(1:n() / n())) %>%
  ungroup()

# Check if '-log10pval' column exists, if not, add it
if (!('-log10pval' %in% colnames(df))) {
  df <- df %>%
    mutate(`-log10pval` = -log10(pval))
}


```

```{r}
subsample_data = TRUE

colors_qq = c(colors, 'DeepRVAT-Synonymous' = 'violet')
```

```{r}
#sample data to make plotting faster
sample_data <- function(df) {
  df %>%
    group_by(Method) %>%
    mutate(sample_percent = ifelse(`-log10pval_expected` < 2, 0.10, 0.5)) %>%
    filter(runif(n()) < sample_percent)
}

# Sample the data --> Comment this in if you want to make the qqplot plotting faster
if (subsample_data){
  df_sub <- df %>%
    sample_data()
}else{
  df_sub = df
}
nrow(df)
#show all values > max_val as triangles
max_val_binary = 30
max_val_quantitative = 50

df_sub <- df_sub %>%
  mutate(
    shape = case_when(
      trait_group == 'Binary' &  `-log10pval` > max_val_binary ~ 'triangle',
      trait_group == 'Binary' &  `-log10pval` <= max_val_binary ~ 'circle',
      trait_group == 'Quantitative' &  `-log10pval` > max_val_quantitative ~ 'triangle',
      trait_group == 'Quantitative' &  `-log10pval` <= max_val_quantitative ~ 'circle',
      TRUE ~ NA_character_
    ),
     `-log10pval` = case_when(
      (trait_group == 'Binary' &  `-log10pval` > max_val_binary) |
      (trait_group == 'Quantitative' &  `-log10pval` > max_val_quantitative) ~ max_val_binary,
      TRUE ~  `-log10pval`
    )
  ) %>%
  mutate(alpha = ifelse(`-log10pval_expected` < 4, 'alpha_low', 'alpha_high'))
 qq_p = ggplot(df_sub %>% filter(Method !=  "DeepRVAT-Synonymous"), 
               aes(x = `-log10pval_expected`, y = `-log10pval`, color = Method, fill = Method, shape = shape, size = shape)) + #, alpha = `-log10pval_expected`)) +
      labs(x = TeX("$Expected -log_{10}(P)$"), y = TeX("$Observed -log_{10}(P)$")) +
      # facet_wrap(vars(trait_group), scales = 'free') +
      geom_abline(intercept = 0, slope = 1, color = 'darkgrey') +
      geom_point(alpha = 0.6) +
      ylim(0, 30) +
      theme_classic() +
      base_theme_wo_margin +
      theme(legend.position = 'none',#,
            ) +
      scale_color_manual(values = colors_qq) +
      scale_fill_manual(values = colors_qq) +
      scale_shape_manual(values = c('triangle' = 17,'circle' = 19)) +
      scale_size_manual(values = c('triangle' = 1.5,'circle' = 0.5))  


```

```{r}
qq_p_syn = ggplot(df_sub %>% filter(grepl('DeepRVAT', Method)), 
               aes(x = `-log10pval_expected`, y = `-log10pval`, color = Method, fill = Method, shape = shape, size = shape)) + #, alpha = `-log10pval_expected`)) +
      labs(x = TeX("$Expected -log_{10}(P)$"), y = TeX("$Observed -log_{10}(P)$")) +
      # facet_wrap(vars(trait_group), scales = 'free') +
      geom_abline(intercept = 0, slope = 1, color = 'darkgrey') +
      geom_point(alpha = 0.6) +
      ylim(0, 30) +
      theme_classic() +
      base_theme_wo_margin +
      theme(legend.position = 'none',#,
            ) +
      scale_color_manual(values = colors_qq) +
      scale_fill_manual(values = colors_qq) +
      scale_shape_manual(values = c('triangle' = 17,'circle' = 19), guide = 'none') +
      scale_size_manual(values = c('triangle' = 1.5,'circle' = 0.5), guide = 'none')  

qq_p_syn
```



# Generate all plots

```{r}
trait_title_dict = c('Training Trait' = 'Traits used in gene impairment module training', 'New Trait' = 'Traits not used in gene impairment module training')
for (t_group in c('Training Trait', 'New Trait')){
  p = ggplot(
        grouped_counts %>% filter(pheno_group == t_group & Method %in% main_methods),
        aes(
          x=trait_with_count,
          y=n,
          fill=Method,
        )
    ) + 
    geom_col(position='dodge', width=0.9, color='darkgray', linewidth=0.1) +
    labs(y = "Significant gene-trait associations", title = trait_title_dict[[t_group]]) +
    theme_classic() +
    # facet_grid(.~is_single_trait, scales='free', space = 'free') + #,  space = "free_x") +
    ggforce::facet_row(vars(is_single_trait), scales = 'free', space = 'free') +
    base_theme_wo_strip +
      theme(
        axis.text.x=element_text(angle=45, vjust=1, hjust=1),
        plot.title = element_text(hjust = 0.5, size = title_font_size),
        axis.title.x=element_blank(),
        legend.key.size=unit(0.5, "line"),
        legend.text = element_text(size = 12),
        panel.spacing=unit(0.75, "cm"),
         strip.background = element_blank(),
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) 
  if (t_group == 'Training Trait'){
    p = p + 
      scale_fill_manual(values = colors,  guide = guide_legend(
      direction = "horizontal", nrow = 2
    )) + theme(legend.position = c(0.65, 0.8)) 
  }else{
    p = p + theme(legend.position = 'none') +
       scale_fill_manual(values = colors) 
  }
  
  print(p)
  plot_list[[paste('quant_discoveries', gsub(' ', '_', t_group), sep = "_")]] = p
}
grouped_counts %>% filter(trait_group == 'All Traits')
```

## timing plot
```{r}
this_df_x_fold = readRDS(file.path(timing_data_dir, paste0('timing_data_x_fold_34_pheno.Rds'))) #generate using deeprvat-analysis/timing/timing.Rmd

plot_colors_deeprvat = c('Monti' = "aquamarine4", 'STAAR' = '#E6AB02','DeepRVAT (with training)' = '#7A2961', 'DeepRVAT (pre-trained)' = '#7A2961')
plot_rename = c('Monti' = "Monti", 'STAAR' = 'STAAR','DeepRVAT (with training)' = 'DeepRVAT\n(with training)', 'DeepRVAT (pre-trained)' = 'DeepRVAT\n(pre-trained)')
days = c(1,  4,  16, 32, 64, 128)
days_sec = days * (3600*24)

 p_pheno_x_fold = ggplot(this_df_x_fold %>% filter(mode == 'time_test' ) ,
       aes(x = Method, y = time, color = Method, fill = Method)) +
  geom_bar(stat="identity") +
  geom_line() +
  scale_color_manual(values = plot_colors_deeprvat, guide = 'none') + 
  scale_fill_manual(values = plot_colors_deeprvat, guide = 'none') + 
  geom_text(aes(label=time_x_val),vjust=-0.5, size = 5) + 
  labs(y = paste('Testing time for 34 phenotypes [d]'), x = '', color = '') +
  scale_y_continuous(breaks = days_sec, labels = paste0(days, 'd'), expand = expansion(mult = c(0,0.3)))+ #, limits = c(0, max(days_sec))) +
  theme_classic() +
  base_theme_wo_margin +
  scale_x_discrete(breaks = names(plot_rename), labels = plot_rename) +
  theme(
        axis.text.x = element_text(angle = 45,  hjust =1, size = font_size)
        )

p_pheno_x_fold

```

## deeprvat score vs phenotype

data generated in ../deeprvat_score_analysis/analyse_500k_burdens_new.Rmd
```{r}
scatter_data = read_parquet(scatter_data_dir)
burden_renamer = c('deeprvat' = 'DeepRVAT', 'alphamissense' = 'AlphaMissense')
remove_pheno_outliers = TRUE

pheno_oi = "Vitamin_D"
gene_oi = "1225"


pheno_oi = "LDL_direct_statin_corrected"
gene_oi = "9138" #gene_names_id_dict[["LDLR"]]

p_df = scatter_data %>% filter(burden_type == "deeprvat")%>%
  filter(gene == gene_oi & phenotype == pheno_oi)

gene_name = unique(p_df[['gene_name']])
  
counts = p_df %>% drop_na() %>%
    group_by(gene, burden_type) %>% 
    summarise(r = cor(gene_score, y), n = n())
    
counts = p_df %>%
  drop_na() %>%
  group_by(gene, burden_type) %>%
  summarise(r = cor.test(gene_score, y)$estimate,
            p_value = cor.test(gene_score, y)$p.value,
            n = n())
  
if(remove_pheno_outliers){
    pheno_vals = p_df[['y']]
    max_pheno_val = quantile(pheno_vals, 0.999, na.rm = TRUE)
    min_pheno_val = quantile(pheno_vals, 1-0.999, na.rm = TRUE)
    p_df = p_df %>% filter(y < max_pheno_val & y > min_pheno_val)
}

p_df = p_df %>%
  mutate(burden_type = burden_renamer[burden_type])

p_score_vs_pheno = ggplot(p_df %>% filter(gene_score != 0))+
  geom_point(aes(x = gene_score, y = y, color = factor(has_plof)), alpha = 0.6, size = 0.5) +
  scale_color_manual(values = c('black', 'red'), labels = c('False', 'True')) +
  geom_smooth(data = p_df %>% filter(gene == gene_oi ), aes(x = gene_score, y = y))+ #, method = 'lm') +
  theme_classic() + 
  # theme(aspect.ratio = 1.7) +
  labs(y = gsub("_", " ", gsub("_statin_corrected", "", pheno_oi)) , 
       # y = 'pheno',
       x = 'Gene score',
       color = "Has pLoF",
       title = gene_name) +
  base_theme_wo_margin +
  theme(legend.position = c(0.8, 0.9), 
        legend.title = element_text(size = 10), 
        legend.box.background = element_rect(colour = "black", fill = 'white'),
        plot.title = element_text(face = "italic")) 

p_score_vs_pheno
```


## combine figures




```{r}
width = 12
label_size = 25
label_specs = list(size = label_size, color = "black", face = "bold", family = NULL)

axis_tex_size = 12
font_theme = theme(axis.text = element_text(size = axis_tex_size), title = element_text(size = axis_tex_size), axis.title = element_text(size = axis_tex_size), strip.text = element_text(size = 20),
                   legend.title = element_text(size = axis_tex_size), legend.text = element_text(size = axis_tex_size))
p = ggarrange(ggarrange(plot_list[['quant_discoveries_Training_Trait']] + font_theme + theme(plot.title = element_text(hjust = 0.5, size = axis_tex_size+2), legend.title = element_blank(), legend.text = element_text(size = axis_tex_size + 2)) , 
                        qq_p + theme(plot.margin = margin(1, 1, 3.1, 1, "cm")) + font_theme, 
                        widths= c(0.65, 0.35), labels = c('a', 'b'), font.label = label_specs), 
    ggarrange(plot_list[['replication_training']] + theme(axis.title.x = element_text(vjust = 10)) + theme(plot.margin = margin(0, 2, 0, 0, "cm"))+ font_theme, 
              p_score_vs_pheno + theme(axis.title.x = element_text(vjust = 10)) + theme(plot.margin = margin(0, 1, 0, 1, "cm"))+ font_theme,
              p_pheno_x_fold  + theme(plot.margin = margin(0, 0, 0, 2, "cm")) + font_theme + theme(axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = axis_tex_size)) , 
                          widths= c(0.35,0.3,  0.35), 
                          nrow = 1, ncol =3, labels = c('c', 'd', 'e'), font.label = label_specs, align = 'h'),
            ggarrange(
              plot_list[['quant_discoveries_New_Trait']]+ font_theme + theme(plot.title = element_text(hjust = 0.5, size = axis_tex_size + 2)), 
              plot_list[['replication_new']] + theme(plot.margin = margin(1, 1, 3, 1, "cm"))+ font_theme, 
              widths= c(0.65, 0.35), labels = c('f', 'g'), font.label = label_specs), 
            nrow = 3, ncol = 1, heights = c(0.5, 0.5, 0.55))


ggsave(file.path(plot_dir, 'figure2.png'), p, width = width , height = 1.2*width, dpi = 600) #, dpi = 600)

```



### supplementary figure synonymous variants

```{r}

count_syn_combined = bind_rows(grouped_counts %>% filter(Method == 'DeepRVAT' & trait_group == "All Traits" & pheno_group == 'Training Trait') %>% select(-is_single_trait), 
          counts_deeprvat_quant_syn %>% filter(Method == 'DeepRVAT' & Trait == 'All traits') %>% select(-is_single_trait) %>% mutate(Method = "DeepRVAT-Synonymous"))

 count_syn = ggplot(
        count_syn_combined,
        aes(
          x=Method,
          y=n,
          fill=Method,
        )
    ) + 
    geom_col(position='dodge', width=0.9, color='darkgray', linewidth=0.1) +
    labs(y = "Significant gene-trait associations") +
    theme_classic() +
    # facet_grid(.~is_single_trait, scales='free', space = 'free') + #,  space = "free_x") +
    base_theme_wo_strip +
      theme(
        axis.text.x=element_blank(),
        axis.ticks.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size = title_font_size),
        axis.title.x=element_blank(),
        legend.key.size=unit(0.5, "line"),
        legend.text = element_text(size = 12),
        panel.spacing=unit(0.75, "cm"),
         strip.background = element_blank(),
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, .1)))  +
    scale_fill_manual(values = colors_qq) +
   geom_text(aes(label=n, vjust = -1))

 
 count_syn
```




```{r}
p_syn_combined = ggarrange(ggarrange(NA, NA, labels = c('a','b'), font.label = label_specs), 
            ggarrange(qq_p_syn +  theme(legend.text = element_text(size = 12)), 
                      count_syn + theme(legend.text = element_text(size = 12)), common.legend = TRUE, legend = 'bottom'),
          nrow = 2, heights = c(0.1, 0.9)
          ) 

ggsave(file.path(plot_dir, 'supp_fig2_syn_vars.png'), p_syn_combined, width = 8, height = 5, dpi = 600)

```


