```{r}
library(reticulate)
library(dplyr)
library(tidyr)
library(ggpubr)
library(arrow)
library(ggplot2)
library(cowplot)
library(gridExtra)

```

The gene file timing_genes.parquet  comprises 1000 randomly chosen genes. 
For the timing experiments, the number of regression chunks was set to 1 in all snakefiles. 
The timing experiments were only run for a single phenotype (Triglycerides)

```{r}
## define directories here 
timing_dir_staar = 'timing_staar'
timing_dir_monti = 'timing_mont'
timing_dir_seed_genes = 'timing_burden_skat'
complete_seed_gene_experiment_dir = '../figure2_association_testing_benchmark/burden_skat'
deeprvat_train_time_res = 'timing_results.parquet'
deeprvat_timing_dir = 'timing_deeprvat'

```

```{r}
plot_colors = c('Monti' = "aquamarine4", 'STAAR' = '#E6AB02','DeepRVAT' = '#7A2961')
plot_colors_deeprvat = c('Monti' = "aquamarine4", 'STAAR' = '#E6AB02','DeepRVAT (with training)' = '#7A2961', 'DeepRVAT (pre-trained)' = '#7A2961')

facet_background = strip.background = element_rect(fill = "white", colour = "black", 
                linewidth = rel(2))

fancy_scientific <- function(l) {
     l <- format(l, scientific = TRUE)
     l <- gsub("^(.*)e", "'\\1'e", l)
     l <- gsub("e", "%*%10^", l)
     l <- gsub("\\+", "", l)
     parse(text=l)
}
my_margin = margin(t = 20, r = 0, b = 0, l = 10)

```

# Load data and define list of genes

## STAAR
```{r}
vtypes_staar  = c('disruptive_missense', 'missense',  'plof',  'plof_disruptive_missense',  'synonymous')
exclude_synonymous_staar = FALSE
```

```{r}
gene_list_staar = list()
staar_results = list()
for (v_filter in vtypes_staar){
  this_result = read_parquet(file.path(timing_dir_staar, 'Triglycerides', v_filter, 'results/burden_associations_testing.parquet')) %>% distinct() %>%
    mutate(rep = v_filter)
  staar_results[[v_filter]] = this_result
  this_genes = this_result %>% distinct(gene) %>% pull(gene)
  gene_list_staar = append(gene_list_staar, list(this_genes))
}

genes_staar = Reduce(intersect, gene_list_staar)
staar_results = do.call(rbind, staar_results) 
if (exclude_synonymous_staar){
  staar_results = staar_results %>% filter(rep != 'synonymous')
}
print(length(genes_staar))
staar_results %>% distinct(rep)

staar_results %>% select(rep, time_inner, time_outer)
```


```{r}
staar_results %>% select(gene, time_outer, time_inner, rep) %>% 
  mutate(time_outer = as.numeric(time_outer),
         time_inner = as.numeric(time_inner)) %>%
  rename('time_test' = 'time_inner', 'time_complete' = 'time_outer') %>%
  filter(gene == 9)
```

```{r}
avg_staar = staar_results %>% select(gene, time_outer, time_inner) %>% 
  mutate(time_outer = as.numeric(time_outer),
         time_inner = as.numeric(time_inner)) %>%
  rename('time_test' = 'time_inner', 'time_complete' = 'time_outer') %>%
  group_by(gene) %>% summarise_each(funs(sum(., na.rm = TRUE)))

avg_staar
```
## monti
```{r}
vtypes_monti  = c('deepripe','missense', 'missense_plof', 'plof', 
                  'spliceai',
                  'spliceai_plof')
```


```{r}
gene_list_monti = list()
monti_results = list()
for (t_type in c('burden', 'skat')){
  for (v_filter in vtypes_monti){
    pickle_output = py_load_object(file.path(timing_dir_monti, 'Triglycerides', v_filter, t_type, 'results/timing_chunk_0.pickle'))
    print(paste(t_type, v_filter, length(pickle_output)))
    t1 = tibble(gene = as.integer(names(pickle_output[[1]])), time_test = as.numeric(pickle_output[[1]]))
    t2 = tibble(gene = as.integer(names(pickle_output[[2]])), time_complete = as.numeric(pickle_output[[2]]))
    this_result = left_join(t1, t2) %>%
      mutate(rep = paste(t_type, v_filter, sep = '-'))
    monti_results[[paste(t_type, v_filter, sep = '-')]] = this_result
    this_genes = this_result %>% distinct(gene) %>% pull(gene)
    gene_list_monti = append(gene_list_monti, list(this_genes))
  }
}
names(monti_results)
genes_monti = Reduce(intersect, gene_list_monti)
monti_results = do.call(rbind, monti_results)
print(length(genes_monti))
monti_results

```


```{r}
monti_filters_to_exclude = c('burden-deepripe', 'skat-plof') #tests not used by Monti
avg_monti = monti_results %>% 
  filter(!grepl(paste(monti_filters_to_exclude, collapse = '|'), rep)) %>%
  select(gene, time_test, time_complete) %>% 
  group_by(gene) %>% summarise_each(funs(sum(., na.rm = TRUE)))

monti_results %>% distinct(rep)

```

## DeepRVAT

### Seed gene discovery
```{r}

vtypes_seed_genes  = c('missense','plof')

```

```{r}
gene_list_seed_genes = list()
seed_genes_results = list()
for (t_type in c('burden', 'skat')){
  for (v_filter in vtypes_seed_genes){
    pickle_output = py_load_object(file.path(timing_dir_seed_genes, 'Triglycerides', v_filter, t_type, 'results/timing_chunk_0.pickle'))
    print(paste(t_type, v_filter, length(pickle_output)))
    t1 = tibble(gene = as.integer(names(pickle_output)), time_test = as.numeric(pickle_output))
    # t2 = tibble(gene = as.integer(names(pickle_output[[2]])), time_complete = as.numeric(pickle_output[[2]]))
    this_result = left_join(t1, t2) %>%
      mutate(rep = paste(t_type, v_filter, sep = '-'))
    seed_genes_results[[paste(t_type, v_filter, sep = '-')]] = this_result
    this_genes = this_result %>% distinct(gene) %>% pull(gene)
    gene_list_seed_genes = append(gene_list_seed_genes, list(this_genes))
  }
}
names(seed_genes_results)
genes_seed_genes = Reduce(intersect, gene_list_seed_genes)
seed_genes_results = do.call(rbind, seed_genes_results)
print(length(genes_seed_genes))
seed_genes_results

head(gene_list_seed_genes[0])

#sum across all tests for each gene
avg_seed_genes = seed_genes_results %>% 
  select(gene, time_test, time_complete) %>% 
  group_by(gene) %>% summarise_each(funs(sum(., na.rm = TRUE)))
avg_seed_genes
```






```{r}
### get number of considered genes per variant type after EAF filtering 
all_counts = tibble()
for (phenotype in list.dirs(complete_seed_gene_experiment_dir, recursive = FALSE, full.names = FALSE)){
  if (!grepl(".snakemake|logs", phenotype)){ 
    file_path = file.path(complete_seed_gene_experiment_dir, phenotype,'eval/burden_associations.parquet')
    tryCatch(
    {
      this_df <- read_parquet(file_path) %>%
        select(gene, method) %>% 
        distinct() %>%
        group_by(method) %>%
        summarise(n = n()) %>%
        mutate(phenotype = phenotype)
    },
    error = function(e) {
      cat("Warning: The file", file_path, "does not exist.\n")
    }
  )
    all_counts = rbind(all_counts, this_df)
  }
}
mean_n_seed_gene_tests = all_counts %>% group_by(method) %>% summarise(mean_n_genes = mean(n))
mean_n_seed_gene_tests
n_tests_seed_genes = sum(mean_n_seed_gene_tests[['mean_n_genes']])
n_tests_seed_genes
this_df <- read_parquet(file_path) 
```




### DeepRVAT
```{r}
repeats = seq(0,5,1)

gene_list_deeprvat = list()
deeprvat_results = list()

for (this_repeat in repeats){
  this_result = read_parquet(file.path(deeprvat_timing_dir, paste0('repeat_', this_repeat), 'burden_associations_testing.parquet')) %>%
    mutate(rep = this_repeat)
  deeprvat_results[[paste0('repeat_', this_repeat)]] = this_result
  this_genes = this_result %>% distinct(gene) %>% pull(gene)
  gene_list_deeprvat = append(gene_list_deeprvat, list(this_genes))
}

genes_deeprvat = Reduce(intersect, gene_list_deeprvat)
print(length(genes_deeprvat))
deeprvat_results = do.call(rbind, deeprvat_results)
length(intersect(genes_staar, genes_deeprvat))
```

```{r}
deeprvat_avg = deeprvat_results %>% select(gene, full_time, score_time) %>% 
  rename('time_test' = 'score_time', 'time_complete' = 'full_time') %>%
  group_by(gene) %>% summarise_each(funs(sum(., na.rm = TRUE)))
deeprvat_avg
```

# Evaluate results

```{r}
training_time_deeprvat <- read_parquet(deeprvat_train_time_res)
training_time_deeprvat
```


n_cv_splits = 5
n_repeats = 6 # number of ensembled DeepRVAT models per split

#Training and burden computation were retrieved manually from log files of training jobs and burden computation jobs 
deeprvat_train_time = training_time_deeprvat %>% filter(Individuals == 130000) %>% pull(`Training time (s)`)  * n_repeats * n_cv_splits#
deeprvat_burden_time_with_overhead = 250 
deeprvat_burden_time = 233 
```


```{r}
genes_to_keep = Reduce(intersect, list(genes_staar, genes_deeprvat, genes_monti)) #genes_monti
print(length(genes_to_keep))
```
```{r}
n_training_phenotypes = 21

gene_avg_staar = colMeans(avg_staar %>% filter(gene %in% genes_to_keep) %>%
                          select(-gene))
print('STAAR')
gene_avg_staar

gene_avg_monti = colMeans(avg_monti %>% filter(gene %in% genes_to_keep) %>% select(-gene))

print('Monti')
gene_avg_monti

gene_avg_deeprvat = colMeans(deeprvat_avg %>% filter(gene %in% genes_to_keep) %>% select(-gene))

print('DeepRVAT')
gene_avg_deeprvat

gene_avg_seed_genes = colMeans(avg_seed_genes %>% filter(gene %in% genes_to_keep) %>% select(-gene))
print('Baseline Burden/SKAT')
gene_avg_seed_genes
seed_gene_discovery_total_time =  n_tests_seed_genes * gene_avg_seed_genes * n_training_phenotypes
seed_gene_discovery_total_time
```


```{r}
modes = c('time_complete', 'time_test')
mode = 'time_test' 
n_genes = 100


getDeepRvatTimes = function(gene_avg_deeprvat, mode, n_genes){
  gene_time = gene_avg_deeprvat[[mode]]
  deeprvat_times = list()
  for (overhead_suffix in  c('', '_with_overhead')){
    time_with_train = seed_gene_discovery_total_time[[mode]]  + deeprvat_train_time + get(paste0('deeprvat_burden_time', overhead_suffix)) + n_genes * gene_time + gene_avg_seed_genes[[mode]] * n_genes
    # time_without_train =  get(paste0('deeprvat_burden_time', overhead_suffix)) + n_genes * gene_time  
    time_without_train =  get(paste0('deeprvat_burden_time', overhead_suffix)) + n_genes * gene_time  + gene_avg_seed_genes[[mode]] * n_genes
    deeprvat_times[[paste0('with_train', overhead_suffix)]] = time_with_train
    deeprvat_times[[paste0('without_train', overhead_suffix)]] = time_without_train
  }
  res_df = tibble(time = as.numeric(deeprvat_times), 
                  n_genes = n_genes, mode = mode, 
                  Method = 'DeepRVAT',
                  deeprvat_type = names(deeprvat_times))
  return(res_df)
}
getTimeComparison = function(gene_avg, mode, n_genes){
  gene_time = gene_avg[[mode]]
  total_time = gene_time * n_genes
  return(tibble(time = total_time, n_genes = n_genes, mode = mode))
}

```

```{r}
all_times = tibble()
for (n_genes in seq(0, 150000, 100)){
  for (mode in  c('time_complete', 'time_test')){
    deeprvat_times = getDeepRvatTimes(gene_avg_deeprvat, mode, n_genes)
    deeprvat_types = c(deeprvat_times$deeprvat_type)
    all_times = rbind(all_times, deeprvat_times)
    
    monti_time = getTimeComparison(gene_avg_monti, mode, n_genes) %>%
      mutate(Method = 'Monti')
    
    staar_time = getTimeComparison(gene_avg_staar, mode, n_genes) %>%
      mutate(Method = 'STAAR')
    for (type in deeprvat_types){
        all_times = rbind(all_times, monti_time %>% mutate(deeprvat_type = type))
        all_times = rbind(all_times, staar_time %>% mutate(deeprvat_type = type))

    }
  }
}
all_times
deeprvat_times
```

```{r}
df = all_times %>% filter(n_genes == 0) %>% head(6)
df <- df %>% separate(Method, into = c("Method", "deeprvat_type"), sep = "_", extra = 'merge')
deeprvat_types = unique(df$deeprvat_type)
deeprvat_types = deeprvat_types[!is.na(deeprvat_types)]
deeprvat_types
```



```{r}

deeprvat_type_names = c(
  'with_train' = 'With DeepRVAT training without Overhead',
  'without_train' = 'DeepRVAT Pre-trained without Overhead',
  'with_train_with_overhead' = 'DeepRVAT (with training)',
  'without_train_with_overhead' = 'DeepRVAT (pre-trained)'
)

deeprvat_type_names_labeller <- function(variable,value){
  return(deeprvat_type_names[value])
}
deeprvat_name_renamer = function(name){
  if(grepl('DeepRVAT', name)){
    name = ifelse(grepl('with_train', name), 'DeepRVAT (with training)', 'DeepRVAT (pre-trained)')
  }
  return(name)
}

plot_df  = all_times %>% filter(grepl('overhead', deeprvat_type) & mode == 'time_test' )
plot_df = plot_df %>% mutate(Method = ifelse(Method == 'DeepRVAT', paste(Method, deeprvat_type, sep = '-'), Method)) %>%
  select(-deeprvat_type) %>%
  distinct()

plot_df[['Method']] = sapply(plot_df[['Method']], deeprvat_name_renamer)
plot_df = plot_df %>% filter(n_genes < 150000 & time < 2e5)
plot_df = plot_df %>% mutate(line_type = ifelse(Method != 'DeepRVAT (with training)', 'DeepRVAT (pre-trained)', 'DeepRVAT (with training)')) %>% 
  mutate(Method = ifelse(grepl('DeepRVAT', Method), 'DeepRVAT', Method))
```


```{r}
p_timing = ggplot(plot_df %>% filter(n_genes < 150000 & time < 2e5) ,
       aes(x = n_genes, y = time, color = Method, linetype = line_type)) +
  # facet_grid(cols = vars(deeprvat_type),
  #            scales = 'free', labeller = deeprvat_type_names_labeller) +
  # facet_wrap(vars(deeprvat_type)) +
  geom_line() +
  scale_y_continuous(labels = fancy_scientific) +
  scale_color_manual(values = plot_colors) + 
  labs(y = 'Testing time [s]', x = 'Tests', color = '', linetype = '') +
  theme_cowplot() +
  theme(strip.background = facet_background,
        axis.text.x = element_text(angle = 44,  hjust =1))+
        # plot.margin = my_margin,
        # aspect.ratio = 1
        # ) +
  scale_linetype_manual(values = c("solid", "dashed"), 
                         guide = guide_legend(override.aes = list(color = plot_colors['DeepRVAT'])))
p_timing
```


# Testing on multiple phenotypes

```{r}
#time per phenotype: (seed_gene_discovery + train_time) + deeprvat_gene_time * n_genes * n_phenotypes + seed_gene_avg_gene_time * n_genes * n_phenotypes

getDeepRvatTimesMultiPheno = function(gene_avg_deeprvat, mode, n_genes, n_phenotypes = 100){
  # training time with training for deeprvat: 1x training_time + 1x seed_gene_time + 1x burden_time + n_phenotypes * n_genes * gene_time_deeprvat +  n_phenotypes * n_genes * gene_time_seed
  # training time without training for deeprvat: 1x burden_time + n_phenotypes * n_genes * gene_time_deeprvat + n_phenotypes * n_genes * gene_time_seed
  gene_time = gene_avg_deeprvat[[mode]]
  deeprvat_times = list()
  for (overhead_suffix in  c('', '_with_overhead')){
    time_with_train = seed_gene_discovery_total_time[[mode]] + deeprvat_train_time + get(paste0('deeprvat_burden_time', overhead_suffix)) + n_genes * gene_time * n_phenotypes  + gene_avg_seed_genes[[mode]] * n_genes * n_phenotypes
    # time_without_train =  get(paste0('deeprvat_burden_time', overhead_suffix)) + n_genes * gene_time * n_phenotypes
    time_without_train =  get(paste0('deeprvat_burden_time', overhead_suffix)) + n_genes * gene_time * n_phenotypes + gene_avg_seed_genes[[mode]] * n_genes * n_phenotypes
    deeprvat_times[[paste0('with_train', overhead_suffix)]] = time_with_train
    deeprvat_times[[paste0('without_train', overhead_suffix)]] = time_without_train
  }
  res_df = tibble(time = as.numeric(deeprvat_times), 
                  n_genes = n_genes, mode = mode, 
                  Method = 'DeepRVAT',
                  deeprvat_type = names(deeprvat_times))
  return(res_df)
}
getTimeComparisonMultipheno = function(gene_avg, mode, n_genes, n_phenotypes = 100){
  gene_time = gene_avg[[mode]]
  total_time = gene_time * n_genes * n_phenotypes
  return(tibble(time = total_time, n_genes = n_genes, mode = mode))
}

```

```{r}
all_times_multipheno = tibble()
n_genes = 20000 # considers 20k protein coding genes
n_phenotypes = 34
for (mode in  c('time_complete', 'time_test')){
  deeprvat_times = getDeepRvatTimesMultiPheno(gene_avg_deeprvat, mode, n_genes, n_phenotypes)
  deeprvat_types = c(deeprvat_times$deeprvat_type)
  all_times_multipheno = rbind(all_times_multipheno, deeprvat_times)
  
  monti_time = getTimeComparisonMultipheno(gene_avg_monti, mode, n_genes, n_phenotypes) %>%
    mutate(Method = 'Monti')
  
  staar_time = getTimeComparisonMultipheno(gene_avg_staar, mode, n_genes, n_phenotypes) %>%
    mutate(Method = 'STAAR')
  all_times_multipheno = rbind(all_times_multipheno, monti_time %>% mutate(deeprvat_type = type))
  all_times_multipheno = rbind(all_times_multipheno, staar_time %>% mutate(deeprvat_type = type))
}


this_df = all_times_multipheno %>% filter(grepl('overhead', deeprvat_type) &  mode == 'time_test') %>%
  mutate(Method = ifelse(Method == 'DeepRVAT', paste(Method, deeprvat_type, sep = '-'), Method)) %>%
  select(-deeprvat_type) %>%
  distinct() 

this_df[['Method']] = sapply(this_df[['Method']], deeprvat_name_renamer)


this_df_x_fold = this_df %>%
  mutate(ref_value = this_df[['time']][this_df[['Method']] == 'DeepRVAT (pre-trained)']) %>%
  mutate(time_x = time/ref_value, time_h = round(time/(3600*24), 1)) %>%
  mutate(time_x_val = paste0(round(time_x,0), 'x \n (', time_h, ' d)' ))
this_df_x_fold

```



```{r}
n_phenotypes = 34

all_times_multipheno = tibble()
n_genes = 20000 # considers 20k protein coding genes
for (mode in  c('time_complete', 'time_test')){
  deeprvat_times = getDeepRvatTimesMultiPheno(gene_avg_deeprvat, mode, n_genes, n_phenotypes)
  deeprvat_types = c(deeprvat_times$deeprvat_type)
  all_times_multipheno = rbind(all_times_multipheno, deeprvat_times)

  monti_time = getTimeComparisonMultipheno(gene_avg_monti, mode, n_genes, n_phenotypes) %>%
    mutate(Method = 'Monti')

  staar_time = getTimeComparisonMultipheno(gene_avg_staar, mode, n_genes, n_phenotypes) %>%
    mutate(Method = 'STAAR')
  all_times_multipheno = rbind(all_times_multipheno, monti_time %>% mutate(deeprvat_type = type))
  all_times_multipheno = rbind(all_times_multipheno, staar_time %>% mutate(deeprvat_type = type))
}
this_df = all_times_multipheno %>% filter(grepl('overhead', deeprvat_type) &  mode == 'time_test') %>%
  mutate(Method = ifelse(Method == 'DeepRVAT', paste(Method, deeprvat_type, sep = '-'), Method)) %>%
  select(-deeprvat_type) %>%
  distinct()

this_df[['Method']] = sapply(this_df[['Method']], deeprvat_name_renamer)


this_df_x_fold = this_df %>%
  mutate(ref_value = this_df[['time']][this_df[['Method']] == 'DeepRVAT (pre-trained)']) %>%
  mutate(time_x = time/ref_value, time_d = round(time/(3600*24), 1)) %>%
  mutate(time_x_val = paste0(round(time_x,0), 'x \n (', time_d, ' d)' ))
this_df_x_fold

saveRDS(this_df_x_fold,'timing_data_x_fold_34_pheno.Rds') #to use for plotting of the main figure

```

```{r}

days = c(1,  4, 8, 16, 32, 64, 128)
days_sec = days * (3600*24)
p_pheno_x_fold = ggplot(this_df_x_fold %>% filter(mode == 'time_test' ) ,
       aes(x = Method, y = time, color = Method, fill = Method)) +
  geom_bar(stat="identity") +
  geom_line() +
  scale_color_manual(values = plot_colors_deeprvat, guide = 'none') + 
  scale_fill_manual(values = plot_colors_deeprvat, guide = 'none') + 
  geom_text(aes(label=time_x_val),vjust=-1, size = 5) + 
  labs(y = 'Testing time per 34 phenotypes [s]', x = '', color = '') +
  scale_y_continuous(breaks = days_sec, labels = paste0(days, 'd'), expand = expansion(mult = c(0,0.3)))+ #, limits = c(0, max(days_sec))) +
  theme_cowplot() +
  theme(strip.background = facet_background,
        # axis.text.x = element_text(angle = 44,  hjust =1),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.margin = my_margin)

p_pheno_x_fold
```



## Plot training times for different number of individuals



```{r}
p_training_time = ggplot(training_time_deeprvat, aes(x = Individuals, y =`Training time (s)` )) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_cowplot() +
  labs(y = 'Training time [s]', x = 'Individuals') #+
  # theme(aspect.ratio = 1,plot.margin = my_margin)
p_training_time
```




```{r}
timing_supp = ggarrange(p_training_time, p_timing, nrow = 1, align = 'h',
          widths = c(0.4, 0.6), labels = c('a', 'b'), 
          font.label = list(size = 20, color = "black", face = "bold", family = NULL)
)

# plot_out_dir = '.'

# ggsave(file.path(plot_out_dir, 'timing_supp_figure.png'), timing_supp, dpi = 600, width = 10, height = 4)
```




