---
title: "DeepRVAT- Figure 3: Phenotype prediction"
author: "Eva Holtkamp"
output: html_document
---

```{r}
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(cowplot)
library(stringr)
library(arrow)
library(yardstick)
library(gdata)
library(dplyr)
library(tidyr)
library(purrr)
library(latex2exp)

options(dplyr.summarise.inform = FALSE)

```

```{r}
source(file.path('../phenotypes.R'))
```


```{r}
exp_dir = 'phenotype_prediction_exp/' #directory where 
burden_heritability_download_dir = "" # path were Supp. table from Weiner et al. has been downloaded (needed for supp figure.  https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-022-05684-z/MediaObjects/41586_2022_5684_MOESM4_ESM.xlsx)

```

```{r}
source('figure3_paths.R')
source('figure_paths.R')
```



```{r}
font_size <- 10
title_font_size = 12
font_family <- "Helvetica"
plot_font <- element_text(size = font_size, family = font_family)

base_theme <- theme(
    text = element_text(size = font_size, family = font_family),
    axis.text = element_text(size = font_size, family = font_family),
    axis.title = element_text(size = font_size + 2, family = font_family),

    strip.text.x = element_text(size = font_size + 2, family = font_family),
    strip.text.y = element_text(size = font_size + 2, family = font_family),
    
    legend.title=element_blank(),
    panel.background = element_rect(fill='transparent'), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA, linewidth = 0), #transparent plot bg
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent', color=NA), #transparent legend panel,
)
base_theme_wo_strip <- theme(
    text = element_text(size = font_size, family = font_family),
    axis.text = element_text(size = font_size, family = font_family),
    axis.title = element_text(size = font_size + 2, family = font_family),

    strip.text.x = element_blank(),
    strip.text.y = element_blank(),
    legend.title=element_blank(),
    panel.background = element_rect(fill='transparent'), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA, linewidth = 0), #transparent plot bg
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent', color=NA), #transparent legend panel
)
```

```{r}
paper_plots = list()
```



```{r}
linear_model_res_path = file.path(exp_dir, 'models/linear_models')
phenotype_suffix = 'qt'
gene_list_renamer = c('Burden/SKAT combined', 'DeepRVAT discoveries')
names(gene_list_renamer) = c('baseline_only', 'deeprvat_discoveries')

```

```{r}

all_combined_metrics = readRDS(file.path(linear_model_res_path, 'plotting_data', paste0('all_recomputed_metrics_test_', phenotype_suffix,'.Rds'))) 
all_combined_metrics %>% distinct(model_name)
```

```{r}
baseline_btypes = as.character(all_combined_metrics %>% distinct(model_name) %>% pull(model_name))
baseline_btypes =  unique(sapply(strsplit(baseline_btypes, "-"), function(x) x[1]))
baseline_btypes = grep('covariates|deeprvat', baseline_btypes, value = TRUE, invert = TRUE)

method_gene_list_filter = quo((`rare_burden` %in% baseline_btypes & gene_list == 'baseline_only') | (`rare_burden` == 'deeprvat' & gene_list == 'deeprvat_discoveries'))
method_gene_list_filter
```


## Plotting 
```{r}
label_default_size = 14 
default_plot_width = 15.5 


method_names = c("cadd" = 'CADD',
                 "plof" = 'pLOF',
                 "polyphen" = 'PolyPhen2',
                "primateai" = 'PrimateAI',
                "sift"= 'SIFT',
                "splicai" = 'SpliceAI',
                'deeprvat' = 'DeepRVAT',
                'covariates-PRS' = 'common PRS',
                'alphamissense' = 'AlphaMissense',
                'absplice' = 'AbSplice')

viridis_colors = viridis::viridis(n = length(baseline_btypes) + 4)
viridis_colors  = viridis::viridis(n = length(baseline_btypes))
deeprvat_color = '#7A2961'
common_prs_color = 'darkgoldenrod2'



RmUnderscore <- function(string) {
  sub("_", " ", string)
}
remove_x_underscore = scale_x_discrete(labels = ~ gsub("_", " ", .x)) 
remove_y_expand = scale_y_continuous(labels = scales::percent, expand =  expansion(mult = c(0, 0.1)))

facet_background = strip.background = element_rect(fill = "white", colour = "black", 
                linewidth = rel(2))
```

# Load and prepare data

## linear model results
```{r}
model_order = c("covariates", "covariates-PRS", baseline_btypes, "deeprvat")
model_order

all_combined_metrics = all_combined_metrics %>% 
  mutate(gene_list = ifelse(grepl('covariates', model_name), 'None', str_split(model_name, '-',  simplify = TRUE)[,2]))

#dummy rows for covariates for all gene lists
covariates_res = all_combined_metrics %>% filter(grepl('covariates', model_name))
for (this_gene_list in unique(all_combined_metrics$gene_list)){
  all_combined_metrics = rbind(all_combined_metrics, covariates_res %>% mutate(gene_list = this_gene_list))
}
covariates_res

all_combined_metrics =  all_combined_metrics %>%
  separate(model_name, c('one', 'two'), sep = '-', remove = FALSE) %>%
  rowwise() %>%
  mutate(model = case_when(
    model_name == 'covariates-PRS' ~ model_name,
    TRUE ~ one
  )) %>%
  ungroup() %>% 
  select(-one, -two) %>%
  filter(!grepl('None', gene_list))  %>%
  select(-top_quantile) %>%
  filter(model %in% model_order)
all_combined_metrics[['model']] =  factor(all_combined_metrics[['model']], model_order)

all_combined_metrics %>% distinct(phenotype)
```


```{r}
unique(all_combined_metrics$model)
delta_df = all_combined_metrics %>%
  select(-model_name, - method) %>%
  pivot_wider(names_from = model, values_from = .estimate) %>%
  pivot_longer(all_of(grep('covariates', model_order, invert = TRUE, value = TRUE)),
               names_to = 'rare_burden', values_to = 'r2_rare') %>%
  mutate(delta_r2 = r2_rare - `covariates-PRS`) %>%
  mutate(ratio_r2 = delta_r2/`covariates-PRS` ) # %>%
delta_df %>% filter(.metric == 'rsq')


all_combined_metrics %>% filter(grepl('covariates', model_name))

```

```{r}
method_map =  c('topq' = 'Top', 'bottomq' = 'Bottom', 'both' = 'highest/largest')
```

## logistic model results
```{r}
logistic_model_res_path = file.path(exp_dir, 'models/logistic_models/')


logistic_model_metrics = readRDS(file.path(logistic_model_res_path, 'plotting_data', paste0('combined_metrics_', phenotype_suffix,'.Rds'))) 

logistic_model_metrics = logistic_model_metrics %>% 
  separate(model_name, c('one', 'gene_list'), sep = '-', remove = FALSE) %>%
  rowwise() %>%
  mutate(rare_burden = case_when(
    model_name == 'covariates-PRS' ~ model_name,
    TRUE ~ one
  )) %>%
  ungroup() %>% 
  select(-one) 
logistic_model_metrics %>% distinct(rare_burden)


covariates_res = logistic_model_metrics %>% filter(grepl('covariates', model_name))
for (this_gene_list in na.omit(unique(logistic_model_metrics$gene_list))){
  logistic_model_metrics = rbind(logistic_model_metrics, covariates_res %>% mutate(gene_list = this_gene_list))
}
logistic_model_metrics = drop_na(logistic_model_metrics, gene_list)
rare_burden_models = grep('covariates', unique(logistic_model_metrics$rare_burden), invert = TRUE, value = TRUE)

logistic_model_metrics %>% distinct(rare_burden)

delta_df_logistic = logistic_model_metrics %>% select( -model_name) %>%
  distinct() %>% 
  pivot_wider(names_from = rare_burden, values_from = .estimate) %>%
  pivot_longer(rare_burden_models,
               names_to = 'rare_burden', values_to = 'auprc_rare') %>%
  mutate(delta_auprc = auprc_rare - `covariates-PRS`,
         ratio_delta_auprc = (auprc_rare - `covariates-PRS`)/ `covariates-PRS`)

delta_df_logistic %>% distinct(rare_burden, gene_list)
delta_df_logistic = delta_df_logistic %>% mutate(extreme_quantile_str = paste0(top_quantile * 100, '%'))
delta_df_logistic[['extreme_direction_string']] = method_map[as.character(delta_df_logistic[['extreme_direction']])]
delta_df_logistic = delta_df_logistic %>% mutate(outlier_mode_name = paste(extreme_direction_string, extreme_quantile_str)) 
delta_df_logistic %>% distinct(phenotype)
delta_df_logistic %>% distinct(top_quantile)
```

```{r}
logistic_model_metrics %>% distinct(rare_burden)
```

## Replication in extreme quantiles plot

data generated in in genopheno/phenotype_prediction/replication_in_extremes_plot.R
```{r}

extreme_quantiles = c(0.01)
ranked_df_list = list()
for (outlier_mode in c('topq', 'bottomq')){
  for (extreme_quantile in extreme_quantiles){
      if (outlier_mode == 'both'){
        #here in total extreme_quantile *2 percent have the data been used (exreme_quantile % at each extreme)
        file_path = file.path(linear_model_res_path, 'plotting_data', 'replication_in_extreme', paste0('ranked_df_combined_', 
                                                    outlier_mode, '_', extreme_quantile, '_x2_', phenotype_suffix,'.Rds'))
      }else{
        file_path = file.path(linear_model_res_path, 'plotting_data', 'replication_in_extreme', paste0('ranked_df_combined_', 
                                            outlier_mode, '_', extreme_quantile, '_', phenotype_suffix,'.Rds'))
      }
      ranked_df = readRDS(file_path)
      ranked_df_list = append(ranked_df_list, list(ranked_df))
  }
}

ranked_df_combined = do.call(rbind, ranked_df_list) %>%
  mutate(extreme_quantile_str = paste0(extreme_quantile * 100, '%'))
ranked_df_combined[['outlier_mode']] = factor(method_map[ranked_df_combined[['outlier_mode']]], levels = method_map)

ranked_df_combined_avg = ranked_df_combined %>% group_by(`Rare estimator`, max_rank, extreme_quantile, extreme_quantile_str, outlier_mode) %>%
  summarise(sum = mean(sum)) %>% mutate(phenotype = 'All Traits') %>%
  mutate(outlier_mode_name = paste(outlier_mode, extreme_quantile_str))
ranked_df_combined_avg %>% ungroup() %>% distinct(`Rare estimator`)
```


run genopheno/phenotype_prediction/plot_enrichment_prs_vs_szsore.R
```{r}

enrichment_list = list()
for (this_quantile in c(0.99, 0.999)){
  this_enrichment = readRDS(file.path(linear_model_res_path, 'plotting_data', paste0('enrichment_prs_vs_zscore_', 
                                                                                   this_quantile, '_', phenotype_suffix,'.Rds')))
  enrichment_list = append(enrichment_list, list(this_enrichment))
}
enrichment_combined = do.call(rbind, enrichment_list)


enrichment_combined %>% distinct(phenotype)

```


## enrichment in extreme phenotypes

```{r}
zero_refernce = enrichment_combined %>% filter(min_z_score == 0) %>%
  mutate(prop_zero = extremes/size) %>%
  select(-size, -extremes, -min_z_score)

enrich_score_df = enrichment_combined %>% 
  filter(min_z_score > 0) %>%
  merge(zero_refernce, how = 'left') %>%
  mutate(prop_extreme_samples = extremes/size,
         enrichment = prop_extreme_samples/prop_zero) %>%
  mutate(prs_quantile = paste0(prs_quantile * 100, '% quantile'))

enrich_score_df[!complete.cases(enrich_score_df), ]


avg_rel_improvement_df = enrich_score_df  %>% select(model, prs_quantile, min_z_score, enrichment, phenotype) %>%
  pivot_wider(names_from = model, values_from = enrichment) %>% 
  pivot_longer(cols = c(baseline_btypes, 'deeprvat'), names_to = 'model') %>%
  mutate(improvment_common = value - `covariates-PRS`, rel_improvment_common = improvment_common/`covariates-PRS`) %>%
  group_by(model, prs_quantile, min_z_score) %>%
  summarise(rel_improvment_common = mean(rel_improvment_common), improvment_common = mean(improvment_common)) %>%
  ungroup()

enrich_score_avg_df = enrich_score_df %>% 
  group_by(model_name, model, burden,  prs_quantile, min_z_score) %>%
  summarise(enrichment = mean(enrichment)) %>%
  ungroup()

enrich_score_df %>% filter(min_z_score > 0 & min_z_score < 0.004 & model == 'deeprvat' & prs_quantile == '99% quantile') 
```

### get largest threshold where enrichment for no phenotype/method is 0
```{r}
threshold_range <- seq(1, 4, by = 0.01)

# Initialize the max_z_score variable
max_z_score <- 0

# Iterate over each threshold in the range
for (threshold in threshold_range) {
  filtered_df <- enrich_score_df[enrich_score_df$min_z_score <= threshold, ]
  
  if (all(filtered_df$enrichment != 0)) {
    max_z_score <- threshold
  }
}

# Print the largest threshold at which no value in Value2 is zero
print(max_z_score)
```

## Main figures

```{r}
## determine boxplot order (lowest to highest median from linear models)
method_order = delta_df %>%
             filter(.metric == 'rsq') %>%
               filter(!! method_gene_list_filter) %>%
  group_by(rare_burden) %>%
  summarise(med = median(ratio_r2)) %>% arrange(med) %>%
  pull(rare_burden)

method_order

color_mapping = c(tail(viridis_colors, -4), deeprvat_color, common_prs_color)
color_mapping = c(head(viridis_colors, length(method_order) -1 ), 
                  deeprvat_color, common_prs_color)
names(color_mapping) = c(names(method_names[method_order]), 'covariates-PRS')

color_mapping

color_mapping_sub = color_mapping[grep('_', names(color_mapping), invert = TRUE)]

rare_method_fill = scale_fill_manual(values = color_mapping)
rare_method_color = scale_fill_manual(values = color_mapping)
```

```{r}
# mean and median imporovement
delta_df %>%
             filter(.metric == 'rsq') %>%
               filter(!! method_gene_list_filter) %>% 
  group_by(rare_burden) %>%
  summarise(median = median(ratio_r2)*100, mean = mean(ratio_r2)*100)
```


```{r}
paper_plots = list()
```


```{r}
delta_df %>%
             filter(.metric == 'rsq') %>%
               filter(!! method_gene_list_filter ) %>% 
  arrange(desc(ratio_r2))

delta_df %>%
             filter(.metric == 'rsq') %>%
               filter(!! method_gene_list_filter ) %>% 
  group_by(rare_burden) %>% 
  summarize(median = median(ratio_r2))
```


 
```{r}
compare_burdens = grep('deeprvat', unique(delta_df[['rare_burden']]), invert = TRUE, value = TRUE)

result_df = delta_df %>% filter(.metric == 'rsq' & gene_list == 'baseline_only')

runWilcoxPaired = function(result_df, metric_col,
                           compare_burdens = c("absplice","alphamissense", "cadd","plof","polyphen","primateai","sift","splicai")){
  all_res = tibble()
  for (method in compare_burdens){
    print(method)
    v1 = result_df %>%
      arrange(phenotype) %>%
      filter(rare_burden == 'deeprvat') %>%
      pull(metric_col)
    
    v2 =  result_df %>%
      arrange(phenotype) %>%
      filter(rare_burden == method) %>% 
      pull(metric_col)
    
    t = wilcox.test(v2, v1, alternative = 'less', paired = TRUE)
    res = tibble(b_da = 'deeprvat', b_alt = method, pval = t$p.value)
    all_res = rbind(all_res, res)
    print(t$p.value)
  }
  return(all_res)
}

```


```{r}
compar_lin = runWilcoxPaired(delta_df %>%
             filter(.metric == 'rsq') %>%
               filter(!! method_gene_list_filter), metric_col = 'ratio_r2')

runWilcoxPaired(delta_df %>% filter(.metric == 'rsq' & gene_list == 'deeprvat_discoveries'),metric_col = 'ratio_r2')
runWilcoxPaired(delta_df %>% filter(.metric == 'rsq' & gene_list == 'baseline_only'),metric_col = 'ratio_r2')

arrange(compar_lin, pval)
```


```{r}
n_items = round(length(unique(delta_df[['rare_burden']])) /2, 0)
p_box_lin = ggplot(data=delta_df %>%
             filter(.metric == 'rsq') %>%
               filter(!! method_gene_list_filter)) +
          geom_boxplot(aes(y=ratio_r2, x = rare_burden, color = rare_burden)) +
          # rare_method_fill +
          labs( y= expression(paste("Relative \u2206", R^2)), fill = 'Rare variant \nscoring method', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = c(head(baseline_btypes, n_items), 'deeprvat', tail(baseline_btypes, -n_items)),
                             labels = method_names) +
          scale_fill_manual(values = color_mapping, limits = names(color_mapping),labels = names(color_mapping), guide = 'none') + #limits = c(baseline_btypes, 'deeprvat')
          scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + #limits = c(baseline_btypes, 'deeprvat') +
          theme_cowplot() +
          remove_y_expand +
          theme(axis.text.x = element_text(angle = 44,  hjust =1)) +
          geom_bracket(data = compar_lin, aes(xmin = b_da, xmax = b_alt, label = signif(pval, 2)), y.position = 0.05, step.increase = 0.1)

print(p_box_lin)

# paper_plots[['linear_boxplot']] = p_box_lin
```



```{r}

# only inidicate if DeepRVAT  is not significantly larger than others 
p_box_lin_ns_mark_only = ggplot(data=delta_df %>%
             filter(.metric == 'rsq') %>%
               filter(!! method_gene_list_filter),
           aes(y=ratio_r2, x = rare_burden, color = rare_burden)) +
          geom_boxplot() +
          # rare_method_fill +
          labs( y= expression(paste("Relative \u2206", R^2)), fill = 'Rare variant \nscoring method', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = method_order,
                             labels = method_names) +
          scale_fill_manual(values = color_mapping, limits = names(color_mapping),labels = names(color_mapping), guide = 'none') +
          scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') +
          theme_classic() +
          base_theme_wo_strip +
          remove_y_expand +
          theme(axis.text.x = element_text(angle = 44,  hjust =1)) 
          # everything is siginificant ( p> 0.05) so no brackets needed and plotting gives an error if the data frame es empty 
          # geom_bracket(data = compar_lin %>% filter(pval > 0.05), aes(xmin = g1, xmax = g2, label = signif(pval, 2)), y.position = 0.05, step.increase = 0.1)

print(p_box_lin_ns_mark_only)
paper_plots[['linear_boxplot']] = p_box_lin_ns_mark_only
```






### check effect of gene lists on significance
```{r}
this_top_quantile = 0.01

for (top_bottom_q in c('bottomq', 'topq')){
  print("deeprvat_discoveries")
  res = runWilcoxPaired(delta_df_logistic %>%
     filter(extreme_direction == top_bottom_q & top_quantile == this_top_quantile) %>%
         filter(.metric == 'pr_auc') %>%
           filter(gene_list == 'deeprvat_discoveries'), metric_col = 'ratio_delta_auprc')
  
  print(res %>% filter(pval> 0.05))
  print("baeline discoveries")

  res = runWilcoxPaired(delta_df_logistic %>%
       filter(extreme_direction == top_bottom_q & top_quantile == this_top_quantile) %>%
           filter(.metric == 'pr_auc') %>%
             filter(gene_list == 'baseline_only'), metric_col = 'ratio_delta_auprc')
  print(res %>% filter(pval> 0.05))
    
}
```


## Logistic model
```{r}
this_top_quantile = 0.01
p_box_list = list()
n = 0
for (top_bottom_q in c('bottomq', 'topq')){
  plot_df = delta_df_logistic %>%
     filter(extreme_direction == top_bottom_q & top_quantile == this_top_quantile) %>%
         filter(.metric == 'pr_auc') %>%
           filter(!! method_gene_list_filter)
  
  # method_order = plot_df %>%
  #           group_by(rare_burden) %>%
  #           summarise(med = median(ratio_delta_auprc)) %>% arrange(med) %>%
  #           pull(rare_burden)
  
  test_metrics = runWilcoxPaired(plot_df, metric_col = 'ratio_delta_auprc')
  test_metrics = test_metrics %>% filter(pval > 0.05)
  y_bracket = quantile(plot_df[['ratio_delta_auprc']], 0.996)
  p_box = ggplot(data= plot_df) +
            # geom_boxplot(aes(x = rare_burden, y = ratio_delta_auprc), outliers = FALSE)
            geom_boxplot(aes(x = rare_burden,  y = ratio_delta_auprc, color = rare_burden)) +
            geom_hline(yintercept = 0, color = 'grey60') +
            facet_wrap(vars(outlier_mode_name), nrow = 1) +
          scale_x_discrete(limits = method_order,
                             labels = method_names) +
          scale_fill_manual(values = color_mapping, limits = names(color_mapping),labels = names(color_mapping), guide = 'none') +
          scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') +
          theme_classic() +
          base_theme +
          labs( y= expression(paste("Relative \u2206AUPRC")), 
                fill = 'Rare variant \nscoring method', x = '') + # 'Relative  \u2206  R2'
            scale_y_continuous(labels = scales::percent) +
           theme(axis.text.x = element_text(angle = 44,  hjust =1), strip.background = element_blank())
  if (nrow(test_metrics) > 0){
    p_box =  p_box + geom_bracket(data = test_metrics, aes(xmin = b_da, xmax = b_alt, label = "NS"), 
                                  y.position = y_bracket, step.increase = 0.1, label.size = 3)
  }       
     
  if (n > 0){
    p_box = p_box + theme(axis.title.y = element_blank())
  }
  p_box_list = append(p_box_list, list(p_box))
  n = n +1
}
box_combined = ggarrange(plotlist = p_box_list)
paper_plots[['box_logistic']] = box_combined
box_combined
```



```{r}
delta_df_logistic %>%
     filter(extreme_direction == 'bottomq' & top_quantile == this_top_quantile) %>%
         filter(.metric == 'pr_auc') %>%
           filter(!! method_gene_list_filter) %>% arrange(desc(ratio_delta_auprc))
```

```{r}
delta_df_logistic %>%
     filter(top_quantile == this_top_quantile) %>%
    filter(.metric == 'pr_auc') %>%
           filter(!! method_gene_list_filter) %>% 
  arrange(desc(ratio_delta_auprc))
```




```{r}
n_test_samples = 224817
this_extreme_quantile = '1%'
max_x_rank = n_test_samples * 0.01
outlier_modes_to_keep = c('Top', 'Bottom')

p <- ggplot(data = ranked_df_combined_avg %>% 
  filter(extreme_quantile_str == this_extreme_quantile & outlier_mode %in% outlier_modes_to_keep) %>%
    filter(max_rank <= max_x_rank)#%>% 
    #filter(`Rare estimator` %in% c('plof', 'deeprvat'))
  , aes(x = max_rank, y = sum, color = `Rare estimator`)) +
  # geom_point(size = 0.2) +
  geom_line() +
  facet_wrap(vars(outlier_mode_name),
             scales = 'free')     +
  scale_color_viridis_d() + 
  scale_color_manual(values = color_mapping, guide = 'none') +
  geom_abline(slope = 1, intercept = 0, linetype = 'dotted', color = 'grey') +
  annotate("segment", x = -Inf, xend = Inf, y = -Inf, yend = -Inf) +
  annotate("segment", x = -Inf, xend = -Inf, y = -Inf, yend = Inf) +
  theme_classic() +
  base_theme +
  theme(
        # strip.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_blank()) +
  labs(y = 'Individuals with \n extreme phenotype', x = TeX(r'(Rank ($\Delta_{common - rare~PRS})$)')) #+
  #guides(color = 'none')

p
paper_plots[['extreme_prediction']] = p

```




```{r}
rare_burdens = delta_df %>% distinct(rare_burden) %>% pull
p_legend = ggplot(data=delta_df %>% 
               filter(rare_burden != 'missense') %>%
               filter(gene_list == 'deeprvat_discoveries' & .metric == 'rsq'), 
           aes(x=phenotype, y=delta_r2, fill = rare_burden)) +
          geom_bar(stat="identity", position=position_dodge()) + 
          scale_fill_manual(values = color_mapping[rare_burdens], limits = rare_burdens, labels = method_names[rare_burdens], breaks = rare_burdens) + 
          theme_classic() +
          base_theme_wo_strip +
          theme(axis.text.x = element_text(angle = 44,  hjust =1, size = font_size),
                legend.key.size = unit(0.3, "cm"),
                legend.position="bottom",
                legend.title = element_blank(),
                legend.box.margin = margin(0,0,0,1)) #+
          # guides(fill = guide_legend(nrow = 1))
print(p_legend)
g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

mylegend<-g_legend(p_legend)
mylegend
paper_plots[['legend']] = mylegend
```


```{r}
methods =  avg_rel_improvement_df %>% distinct(model) %>% pull(model)
p = ggplot(data = avg_rel_improvement_df %>% filter(min_z_score <= max_z_score & prs_quantile == '99% quantile'),
         aes(x = min_z_score, y = rel_improvment_common, color = model)) +
  # geom_point(size = 0.2) +
  geom_line(size = 0.3) +
  theme_classic() +
  base_theme_wo_strip +
  scale_color_manual(values = color_mapping[methods], limits = methods,
                     labels = method_names[methods], guide = 'none') +
  #labs(x = 'Phenotype outlier (z-score)', y = 'Relative enrichment vs baseline', color = '', linetype = '') +
  labs(x = 'Phenotype cutoff (minimal z-score)', y = 'Relative enrichment \n for top 1% prediction', color = '', linetype = '') +

  scale_y_continuous(labels = scales::percent) #+


paper_plots[['extreme_enrichment']] = p 
p
avg_rel_improvement_df %>% filter(min_z_score <= max_z_score & min_z_score > max_z_score - 0.005 & prs_quantile == '99% quantile') %>% arrange(desc(rel_improvment_common))
```

```{r}
 ggplot(data = avg_rel_improvement_df %>% filter(min_z_score <= max_z_score & prs_quantile == '99% quantile'),
         aes(x = min_z_score, y = rel_improvment_common, color = model)) +
  # geom_point(size = 0.2) +
  geom_line(size = 0.3) +
  theme_classic() +
  base_theme_wo_strip +
  scale_color_manual(values = color_mapping[methods], limits = methods,
                     labels = method_names[methods]) +
  labs(x = 'Phenotype cutoff (minimal z-score)', y = 'Relative enrichment \n for top 1% prediction', color = '', linetype = '') +
  #labs(x = 'Phenotype outlier (z-score)', y = 'Relative enrichment vs baseline', color = '', linetype = '') +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = c(0.2, 0.5))
```

## Arrange main figure
```{r}
rare_burdens = delta_df %>% distinct(rare_burden) %>% pull
p_legend = ggplot(data=delta_df %>% 
               filter(rare_burden != 'missense') %>%
               filter(gene_list == 'deeprvat_discoveries' & .metric == 'rsq'), 
           aes(x=phenotype, y=delta_r2, fill = rare_burden)) +
          geom_bar(stat="identity", position=position_dodge()) + 
          scale_fill_manual(values = color_mapping[rare_burdens], limits = rare_burdens, labels = method_names[rare_burdens], breaks = rare_burdens) + 
          theme_classic() +
          base_theme_wo_strip +
          theme(axis.text.x = element_text(angle = 44,  hjust =1, size = font_size),
                legend.key.size = unit(0.3, "cm"),
                legend.text = element_text(size = title_font_size),
                legend.position="bottom",
                legend.title = element_blank(),
                legend.box.margin = margin(0,0,0,1)) #+
          # guides(fill = guide_legend(nrow = 1))
print(p_legend)
g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

mylegend<-g_legend(p_legend)
mylegend
paper_plots[['legend']] = mylegend
```

```{r}


(1/6) * 2
lay <- rbind(c(NA,NA,NA,NA,NA,NA,NA,NA,NA),
             c(1,1,1,2,2,2,2,2,2),
             c(NA,3,3,3,3,NA, 5, 5,5),
             c(4,4,4,4,4,NA,5, 5,5))
gt = grid.arrange(paper_plots[['linear_boxplot']],
             paper_plots[['box_logistic']],
            paper_plots[['legend']],
             paper_plots[['extreme_prediction']],
             paper_plots[['extreme_enrichment']] + theme(plot.margin = margin(0, 0.2, 0.9, 0, "cm")),
          nrow = 4, ncol = 9,
          layout_matrix = lay, heights = c(0.1,1.3,0.1,1),
          widths = c(1,1,1,1,1,0.5,1,1,1))
p = as_ggplot(gt) +                                # transform to a ggplot
  draw_plot_label(label = c("a", "b", "c", "d"), size = 20,
                  x = c(0.00, 3/9, 0.05, 5.7/9), y = c(1, 1,0.47, 0.47)) # Add labels

ggsave(file.path(plot_dir, 'figure_4.png'), p, width = 24, height = 20, unit = 'cm')
p
```





# Supplements

## Gene list comparison
```{r}

  

p_box_lin_gene_list = ggplot(data=delta_df %>%
             filter(.metric == 'rsq'),
           aes(y=ratio_r2, x = rare_burden, fill = gene_list)) +
          geom_boxplot() +
          # rare_method_fill +
          labs( y= expression(paste("Relative \u2206", R^2)), fill = 'Gene list', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = method_order,
                             labels = method_names) +
          scale_fill_brewer(palette = 'Pastel1', limits = names(gene_list_renamer), labels = gene_list_renamer) +   
          theme_cowplot() +
          remove_y_expand +
          theme(axis.text.x = element_text(angle = 45,  hjust =1),
                strip.background = facet_background)
print(p_box_lin_gene_list)

```

```{r, eval=FALSE}
delta_df_logistic %>% filter(is.na(metric))
plot_df = delta_df_logistic %>%
   filter(top_quantile == this_top_quantile) %>%
       filter(.metric == 'pr_auc')

p_box_log_gene_list = ggplot(#data=plot_df[plot_df[[metric]] <= y_max,],
               data= plot_df,
           aes_string(y = metric, x = 'rare_burden', fill = 'gene_list')) +
          geom_boxplot() +
          geom_hline(yintercept = 0, color = 'grey60') +
          facet_wrap(vars(outlier_mode_name), nrow = 1) +
          labs( y= expression(paste("Relative \u2206AUPRC")), 
                fill = 'Gene list', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = c(head(baseline_btypes, n_items), 'deeprvat', tail(baseline_btypes, -n_items)),
                           labels = method_names) +
          scale_fill_brewer(palette = 'Pastel1', limits = names(gene_list_renamer), labels = gene_list_renamer) +           # scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + #limits = c(baseline_btypes, 'deeprvat')
          theme_cowplot() +
          scale_y_continuous(labels = scales::percent) +
         theme(axis.text.x = element_text(angle = 44,  hjust =1),
               strip.background = facet_background)
print(p_box_log_gene_list)

```


```{r}
p_box_list_supp = list()
n = 0
metric = "ratio_delta_auprc"
for (top_bottom_q in c('bottomq', 'topq')){
  plot_df = delta_df_logistic %>%
     filter(extreme_direction == top_bottom_q & top_quantile == this_top_quantile) %>%
         filter(.metric == 'pr_auc') %>%
    filter(gene_list != 'PRS')
  
  
  p_box_log_gene_list = ggplot(data=plot_df,
           aes_string(y = metric, x = 'rare_burden', fill = 'gene_list')) +
          geom_boxplot() +
          geom_hline(yintercept = 0, color = 'grey60') +
          facet_wrap(vars(outlier_mode_name), nrow = 1) +
          labs( y= expression(paste("Relative \u2206AUPRC")), 
                fill = 'Gene list', x = '') + # 'Relative  \u2206  R2'
          scale_x_discrete(limits = method_order,
                             labels = method_names) +
          scale_fill_brewer(palette = 'Pastel1', limits = names(gene_list_renamer), labels = gene_list_renamer) +           # scale_color_manual(values = color_mapping, limits = names(color_mapping), guide = 'none') + #limits = c(baseline_btypes, 'deeprvat')
          theme_cowplot() +
          scale_y_continuous(labels = scales::percent) +
         theme(axis.text.x = element_text(angle = 44,  hjust =1),
               strip.background = facet_background) +
    guides(fill = 'none')
  
  if(n > 0){
    p_box_log_gene_list = p_box_log_gene_list + labs(y = '')
  }
  p_box_list_supp = append(p_box_list_supp, list(p_box_log_gene_list))
  n = n + 1
}
p_box_log_gene_list = ggarrange(plotlist  = p_box_list_supp, common.legend = TRUE)

p_box_log_gene_list
```


```{r}
plot_width = 20
p_box_gene_list = ggarrange(p_box_lin_gene_list , 
          p_box_log_gene_list,   common.legend = TRUE,
          widths = c(0.35, 0.65),
          labels = 'auto',
          font.label = list(size = plot_width/default_plot_width * label_default_size),
          legend="bottom") 
p_box_gene_list          
ggsave(file.path(plot_dir, 'figure_4_supp_1.png'), width = plot_width, height = 12, unit = 'cm',
       dpi = 600)
# ggsave(file.path(plot_dir, 'figure_4_supp_1.pdf'), width = plot_width, height = 10, unit = 'cm',
#        dpi = 600)

```









# Compare to burden heritability

```{r}
library(readxl)
library(dplyr)

```




```{r}
burden_her_file = file.path(burden_heritability_download_dir, "41586_2022_5684_MOESM4_ESM.xlsx")
```

```{r}
phecode_dict = c(
 'Apolipoprotein_A'= 30630,
 'Apolipoprotein_B'= 30640,
 'Calcium'= 30680,
 'Cholesterol_statin_corrected'= 30690,
 'HDL_cholesterol'= 30760,
 'IGF_1'= 30770,
 'LDL_direct_statin_corrected'= 30780,
 'Lymphocyte_percentage'= 30180,
 'Mean_corpuscular_volume'= 30040,
 'Mean_platelet_thrombocyte_volume_(MPTV)'= 30100,
 'Mean_reticulocyte_volume'= 30260,
 'Neutrophill_count'= 30140,
 'Platelet_count'= 30080,
 'Platelet_crit'= 30090,
 'Platelet_distribution_width'= 30110,
 'Red_blood_cell_(erythrocyte)_count'= 30010,
 'SHBG'= 30830,
  'Standing_height'= 50,
 'Total_bilirubin'= 30840,
 'Triglycerides'= 30870,
 'Urate'= 30880,
 'Body_mass_index_BMI'= 21001,
 'Glucose'= 30740,
 'Vitamin_D'= 30890,
 'Albumin'= 30600,
 'Total_protein'= 30860,
 'Cystatin_C'= 30720,
 'Gamma_glutamyltransferase'= 30730,
 'Alkaline_phosphatase'= 30610,
 'Creatinine'= 30700,
 'Whole_body_fat_free_mass'= 23101,
 'Forced_expiratory_volume_in_1_second_FEV1'= 20153,
 'Glycated_haemoglobin_HbA1c'= 30750,
 'Mean_platelet_thrombocyte_volume'= 30100,
 'Red_blood_cell_erythrocyte_count'= 30010
)
phecode_dict_df = tibble(phenotype = names(phecode_dict), phenocode = phecode_dict)
```

```{r}
phenotype_key_df <- read_excel(burden_her_file, 
    sheet = "ST5")

burden_herit <- read_excel(burden_her_file, 
    sheet = "ST8")
```

```{r}
burden_herit_deeprvat = phecode_dict_df %>% left_join(phenotype_key_df %>% select(phenocode, phenotype_key)) %>% left_join(burden_herit)
burden_herit_deeprvat
phenotype_key_df %>% filter(phenocode %in% phecode_dict)
```





```{r}
p_df = delta_df %>%
  filter(rare_burden == 'deeprvat' & gene_list == 'deeprvat_discoveries'&`.metric` == 'rsq')%>% 
  left_join(burden_herit_deeprvat) %>% 
  drop_na(aggregated_h2)
cor = round(cor(p_df[['delta_r2']], p_df[['aggregated_h2']], method = 'spearman'), 2)

p_her = ggplot(p_df, aes(y = `delta_r2`, x = aggregated_h2))  + 
  geom_point() +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs( y= expression(paste("\u2206", R^2)), x = 'Burden heritability') +
  annotate("text", y = 0.015, x =0.035, label = sprintf("r = %s", cor), hjust = 1, vjust = 1, size = 5)
p_her  
cor

ggsave(file.path(plot_dir, 'figure_4_supp_2.png'), p_her, width = 10, height = 8, unit = 'cm',
       dpi = 600)
```

## Supp tables

```{r}
this_top_quantile = 0.01 # for logistic models 

col_renamer_linear = c(Trait = "phenotype", "Gene list" = "gene_list", "Rare burden type" = "rare_burden", "Metric" = ".metric", "Metric PRS" =  "covariates-PRS", "Metric rare burden" = "r2_rare", 
  "Delta Metric (rare burden - PRS)" = "delta_r2", 
  "Relative Delta Metric" = "ratio_r2")
res_linear_model = delta_df %>% filter(.metric == 'rsq') %>%
  filter(gene_list %in% names(gene_list_renamer)) %>% 
  select(all_of(col_renamer_linear)) %>% 
  mutate(`Prediction model` = "Linear model")

col_renamer_logistic = c(Trait = "phenotype", "Gene list" = "gene_list","Metric" = ".metric", "Metric PRS" =  "covariates-PRS", "Metric rare burden" = "auprc_rare", 
  "Delta Metric (rare burden - PRS)" = "delta_auprc", 
  "Relative Delta Metric" = "ratio_delta_auprc", "Outlier quantile" = "outlier_mode_name")
res_logistic_model = delta_df_logistic %>% filter(.metric == 'pr_auc' & top_quantile == this_top_quantile) %>%
  filter(gene_list %in% names(gene_list_renamer)) %>% 
  select(all_of(col_renamer_logistic)) %>% 
  mutate(`Prediction model` = "Logistic model")

res_prediction_models = bind_rows(res_linear_model, res_logistic_model)  %>% 
  mutate(`Rare burden type` = method_names[as.character(`Rare burden type`)]) %>%
  mutate(`Gene list` = gene_list_renamer[as.character(`Gene list`)]) %>% 
  mutate(Trait = gsub('_', ' ', Trait)) %>%
  mutate(Trait =  ifelse(Trait %in% names(phenotype_renamer), phenotype_renamer[Trait], Trait))

res_prediction_models
```


```{r}
write.csv(res_prediction_models,
              file.path(supp_table_dir, 'phenotype_prediction_metrics.csv'),
          row.names = FALSE)

```

